---
title: "TMRC3 `r Sys.getenv('VERSION')`: Visualizing Analyses"
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
bibliography: atb.bib
output:
 html_document:
  code_download: true
  code_folding: show
  fig_caption: true
  fig_height: 7
  fig_width: 7
  highlight: zenburn
  keep_md: false
  mode: selfcontained
  number_sections: true
  self_contained: true
  theme: readable
  toc: true
  toc_float:
   collapsed: false
   smooth_scroll: false
---

<style type="text/css">
body .main-container {
  max-width: 1600px;
}
body, td {
  font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
  font-size: 16px
}
</style>

```{r, include=FALSE}
library(broom) ## Provides tidy methods for lm/glm/etc
library(dplyr)
library(forcats)
library(ggplot2)
library(ggstatsplot)
library(glue)
library(hpgltools)
library(lares)

knitr::opts_knit$set(progress = TRUE, verbose = TRUE, width = 90, echo = TRUE)
knitr::opts_chunk$set(
  error = TRUE, fig.width = 8, fig.height = 8, fig.retina = 2,
  out.width = "100%", dev = "png",
  dev.args = list(png = list(type = "cairo-png")))
old_options <- options(digits = 4, stringsAsFactors = FALSE, knitr.duplicate.label = "allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size = 12))
ver <- Sys.getenv("VERSION")
rundate <- format(Sys.Date(), format = "%Y%m%d")

rmd_file <- "02visualization"
savefile <- gsub(pattern = "\\.Rmd", replace="\\.rda\\.xz", x = rmd_file)
loaded <- load(file = glue("rda/tmrc3_data_structures-v{ver}.rda"))
```

# TODO:

1.  Ensure that no variance partition includes biopsies beyond some
    initial cursory.

# Changelog

* Set input data to the new 202212 dataset.  Looking for some messed up colors.
* Reasonably certain I figured out the color discrepency.  I was
  letting the eosinophil dataset choose its own colors rather than
  force them to be the same as the other cell types; even though I
  _thought_ I told them to explicitly set their colors to be the same
  as the others.  I think the changes I made in datasets.Rmd fixed
  this, so I regenerated the rda/etc in that document and am now
  testing the colors here.

# Introduction

Moving all of the visualization and diagnostic tasks to this document.
The metadata and gene annotation data collection tasks are therefore
in tmrc3_data_structures.Rmd.  The reasons for some of the data
structure creation in that document is made clear here.

# Notes

1.  Lesion vs Ulcer: Ulcer is the base of the crater of the lesion
observed.  The lesion is this, the border, and any region with signs
of inflammation.  It is not known if these metrics are equivalent, or
if one is better than the other.  Some people do not have ulcers and
therefore in those cases we can only really consider the lesion size.
E.g. most people in Colombia have ulcers, which are the cratered sore;
however there are a few people who have a 'plaque' or some form of
smaller, less intrusive presentation -- these are still cutaneous.

Thus the lesion size is the more inclusive metric, but potentially
ulcer size is more informative?  Any inflammation in the skin causes
the person to be defined as failure.

2. Note from Maria Adelaida: Some chemokines are suggestive of
   Eosinophil recruitment.

## Goals

These samples are from patients who either successfully cleared a
Leishmania panamensis infection following treatment, or did not.  They
include biopsies from each patient along with purifications for
Monocytes, Neutrophils, and Eosinophils.  When possible, this process
was repeated over three visits; but some patients did not return
for the second or third visit.

The over-arching goal is to look for attributes(most likely genes)
which distinguish patients who do and do not cure the infection after
treatment.  If possible, these will be apparent on the first visit.

```{r}
plot_legend(hs_expt)
plot_nonzero(hs_expt)
```

## Figure S2 + 1: Non-zero genes after sample filtering

The following plot is essentially identical to the previous with two
exceptions:

1.  The samples with too few genes (11,000 currently) are gone.  In
    the current iteration of the datasets Rmd, this comprises either
    two or three samples.
2.  The samples are colored by cure(purple)/fail(yellow)

```{r}
plot_nonzero(tc_valid, plot_labels = FALSE)
```

## Quick picture before removing miltefosine samples

Maria Adelaida's quote: "I would like one picture of all samples
including the miltefosine so that I can keep in my mind why we removed
them."

# PCA with both drugs

The following block will illustrate why we chose to remove the samples
which were treated with miltefosine.  The short reason: too few
samples.  The slightly longer reason: miltefosine has a different mode
of action.

```{r}
tc_expt_norm <- normalize_expt(hs_expt, filter = TRUE, norm = "quant",
                               convert = "cpm", transform = "log2") %>%
  set_expt_batches(fact = "drug")

tc_expt_drug_pca <- plot_pca(tc_expt_norm, cis = NULL)
tc_expt_drug_pca <- plot_pca(tc_expt_norm)
tc_expt_drug_pca

tc_expt_nb <- normalize_expt(hs_expt, filter = TRUE, convert = "cpm",
                             transform = "log2", batch = "svaseq") %>%
  set_expt_batches(fact = "drug")
tc_expt_drug_nb_pca <- plot_pca(tc_expt_nb)
tc_expt_drug_nb_pca

t_expt_drug <- subset_expt(hs_expt, subset = "clinic=='tumaco'")
t_expt_norm <- normalize_expt(t_expt_drug, filter = TRUE, norm = "quant",
                              convert = "cpm", transform = "log2") %>%
  set_expt_batches(fact = "drug")
t_expt_drug_pca <- plot_pca(t_expt_norm)
t_expt_drug_pca

t_expt_nb <- normalize_expt(t_expt_drug, filter = TRUE, convert = "cpm",
                             transform = "log2", batch = "svaseq") %>%
  set_expt_batches(fact = "drug")
t_expt_drug_nb_pca <- plot_pca(t_expt_nb)
t_expt_drug_nb_pca
```

# Host Distributions/Visualizations of interest

The sets of samples used to visualize the data will also comprise the
sets used when later performing the various differential expression
analyses.

## Global metrics

Start out with some initial metrics of all samples.  The most obvious
are plots of the numbers of non-zero genes observed, heatmaps showing
the relative relationships among the samples, the relative library
sizes, and some PCA.  It might be smart to split the library sizes up
across subsets of the data, because they have expanded too far to see
well on a computer screen.

The most likely factors to query when considering the entire dataset
are cure/fail, visit, and cell type.  This is the level at which we
will choose samples to exclude from future analyses.

```{r}
plot_legend(tc_biopsies)
plot_libsize(tc_biopsies)
plot_nonzero(tc_biopsies)
```

plot_libsize_prepost attempts to provide an idea about how much data
is lost when low-count filtering the data.

The first plot it produces is a barplot of the number of reads removed
by the filter from each sample.  The second plot has two bars, the top
bar is labeled with the number of low-count genes before the filter.
The lower bar represents the number after the filter and is assumed to
be quite low.

```{r}
biopsy_prepost <- plot_libsize_prepost(tc_biopsies)
biopsy_prepost
#biopsy_prepost$count_plot
#biopsy_prepost$lowgene_plot
## Minimum number of biopsy genes: ~ 14,000

plot_libsize(tc_eosinophils)
plot_nonzero(tc_eosinophils)
eosinophil_prepost <- plot_libsize_prepost(tc_eosinophils)
eosinophil_prepost[["count_plot"]]
eosinophil_prepost[["lowgene_plot"]]
## Minimum number of eosinophil genes: ~ 13,500

plot_libsize(tc_monocytes)
plot_nonzero(tc_monocytes)
monocyte_prepost <- plot_libsize_prepost(tc_monocytes)
monocyte_prepost[["count_plot"]]
monocyte_prepost[["lowgene_plot"]]
## Minimum number of monocyte genes: ~ 7,500 before setting the minimum.

plot_libsize(tc_neutrophils)
plot_nonzero(tc_neutrophils)
neutrophil_prepost <- plot_libsize_prepost(tc_neutrophils)
neutrophil_prepost[["count_plot"]]
neutrophil_prepost[["lowgene_plot"]]
## Minimum number of neutrophil genes: ~ 10,000 before setting minimum coverage.
```

The above block just repeats the same two plots on a per-celltype
basis: the number of reads observed / sample and a plot of observed
genes with respect to coverage.  I made some comments with my
observations about the number of genes.

# Seeking Confounded/Correlated factors in the metadata

One task we were uncertain of how to address: how best to consider the
many factors provided in the metadata and whether or not they are
hightly correlated or completely confounded.  Theresa provided some
suggestions about how we might measure the degree to which correlated
variables might be a problem and decide which variables we can(not)
include in our statistical models of the data when performing the
differential expression analyses.

## Correlated factors in the Tumaco+Cali data

I am going to implement this in a few steps, first I will do a
cross-correlation of a relatively large array of variables in the
data, then focus on a few which we suspect to be problematic.

```{r}
initial_queries <- c("Sex", "Ethnicity", "Age", "Weight", "Height",
                     "Previously_Diagnosed", "Evolution_Time",
                     "Num_Active_Lesions", "V2_New_Lesions",
                     "V3_New_Lesions", "Adherence", "Therapeutic_Outcome_Final")
initial_df <- demographics_filtered[, initial_queries]
summary(initial_df)
initial_numeric <- initial_df
for (f in colnames(initial_numeric)) {
  initial_numeric[[f]] <- as.numeric(as.factor(initial_numeric[[f]]))
}
initial_cross <- corr_cross(initial_numeric)
pp(file = "images/initial_crosscor.pdf")
initial_cross
dev.off()
initial_cross
```

We should remove height and weight, but I wanted first to see that
everything was working as expected.  We also will want to exclude
final outcome when searching for factors which are unsuitable for
model inclusion.

```{r}
model_test_df <- initial_df
model_test_df[["Weight"]] <- NULL
model_test_df[["Height"]] <- NULL
model_test_df[["Therapeutic_Outcome_Final"]] <- NULL
model_test_numeric <- initial_numeric
model_test_numeric[["Weight"]] <- NULL
model_test_numeric[["Height"]] <- NULL
model_test_numeric[["Therapeutic_Outcome_Final"]] <- NULL

test_cross <- corr_cross(model_test_numeric)
pp(file = "images/model_test_crosscor.pdf")
test_cross
dev.off()
test_cross
```

At this point we have some factors which are flagged as highly
correlated.  Keep this in mind when building models later.

## Regression analyses vs outcome

Now examine a more limited set of likely interesting factors.

!!An important note: 202408!!

In the following block I changed the input from the full merged
demographics by sample (e.g. there are multiple rows for each person
coming from the combination of multiple visits/celltypes), to the one
row/person found in the demographics data.

In addition, in my initial implementation, I did all analyses using
linear regression; Neal kindly pointed out this is not optimal.

Finally, it is probably pretty obvious that this is my first foray
into the usage of regression analyses vis a vis comparing various
metadata factors and estimating their significance with respect to our
outcome variable.  Thus, I kind of fool around in some of the
following blocks.

```{r}
regression_queries <- c("Therapeutic_Outcome_Final", "Weight", "Sex",
                        "Clinic", "Ethnicity", "Age")
regression_df <- demographics_filtered[, regression_queries]
regression_numeric <- regression_df
for (f in colnames(regression_numeric)) {
  regression_numeric[[f]] <- as.numeric(as.factor(regression_numeric[[f]]))
}
cross_df <- regression_df
cross_df[["Therapeutic_Outcome_Final"]] <- NULL
cross_numeric <- regression_numeric
cross_numeric[["Therapeutic_Outcome_Final"]] <- NULL

regression_cross <- corr_cross(cross_df, type = 1)
pp(file = "images/weight_sex_clinic_ethnicity_age_factor_crosscor.pdf")
regression_cross
dev.off()
regression_cross
```

The following is the version which we believe to be the most
appropriate for the reader in a supplemental ~S1

```{r}
regression_cross_numeric <- corr_cross(cross_numeric, type = 1)
pp(file = "figures/weight_sex_clinic_ethnicity_age_numeric_crosscor.svg")
regression_cross_numeric
dev.off()
regression_cross_numeric
```

### Copy these with only the Tumaco people

```{r}
tumaco_idx <- regression_numeric[["Clinic"]] == "2"
t_regression_numeric <- regression_numeric[tumaco_idx, ]
t_regression_df <- regression_df[tumaco_idx, ]
```

Similarly, when we look only at Tumaco, this will also be used in
figure ~S1

```{r}
t_regression_queries <- c("Weight", "Sex", "Ethnicity", "Age")
t_cross_df <- t_regression_numeric[, t_regression_queries]
t_regression_cross <- corr_cross(t_cross_df)
pp(file = "figures/tumaco_weight_sex_ethnicity_age_numeric_crosscor.svg")
t_regression_cross
dev.off()
t_regression_cross
```

Discussion with Maria Adelaida and Neal: 202408

There was a brief discussion regarding how we get to the numeric
correlations in the cross correlation plot.

Najib wants to query the difference between the individual factor
table and the various mixed model regression values.

Why do linear regression vs. logistical regression?

Multilevel regression vs. multiple regression:

multilevel would be used when there is a nested structure to the
experimental design.

multiple regression: applying multiple factors to the regression.

Save confusion by explicitly stating multi-variable.  For the purposes
of this discussion we will avoid any multilevel regression because our
experimental design isn't crazytown.

"The main puzzle": How did sex appear as a strong effect in the
regression when we performed the wilcox test?  It may be that the
model used is inappropriate.

Najib query: when is a mixed effect model appropriate?  lme4 and
multilevel are more closely related and used when there are both fixed
and random effects in the model.  It is likely that if a multilevel
model is not appropriate, then a mixed effect is also not appropriate
(e.g. don't use lmer/lme4).

```{r}
regression_tests <- c("Age", "Clinic", "Ethnicity", "Sex", "Weight")
lm_regression_demographics <- extract_linear_regression(
  regression_numeric, query = "Therapeutic_Outcome_Final", factors = regression_tests,
  excel = glue("excel/numeric_demographics_regression_final_sex_clinic_ethnicity_age-v{ver}.xlsx"))
pp(file = "figures/demographics_only_linear_regression.svg")
lm_regression_demographics[["forest"]]
dev.off()
lm_regression_demographics[["forest"]]
lm_regression_demographics[["summary"]]
```

The following will be used as supplemental figures as well, providing
a delineation between a multi-variable logistic regression and
multiple single variable regressions.

```{r}
log_regression_demographics <- extract_logistic_regression(
  regression_df, query = "Therapeutic_Outcome_Final", factors = regression_tests,
  excel = glue("excel/tc_multivariable_logistic_regression-v{ver}.xlsx"))
pp(file = glue("figures/tc_multivariable_logistic_regression-v{ver}.svg"))
log_regression_demographics[["forest"]]
dev.off()
log_regression_demographics[["forest"]]
log_regression_demographics[["summary"]]

tc_log_iterative_regression_demographics <- iterate_logistic_regression(
  regression_df, query = "Therapeutic_Outcome_Final", factors = regression_tests,
  excel = glue("excel/tc_simple_logistic_regression.xlsx"))
pp(file = glue("figures/tc_simple_logistic_regression.svg"))
tc_log_iterative_regression_demographics[["forest"]]
dev.off()
tc_log_iterative_regression_demographics[["forest"]]
tc_log_iterative_regression_demographics[["summary"]]
```

Discussion of regression with Neal:

The creation of correct models, the forest plots above were created
from a model that looks like ~ 0 + age + sex + clinic + whatever...
Neal is building (I think) toward a conclusion which says that some of
these factors are confounded and may need to be separated.

rule of ten: for each variable in a logistic regression or survival
analysis, one wants to have 10 more entries in the input data.  Thus
we ideally would have 50 cures and 50 fails in order to have this many
factors in the model.  In this data we only have ~ 30 separate people,
so this is not tenable.

I think therefore the most likely thing to build in a plot like this
forest plot would be 1 row each variable using its result from a x ~ y
alone.

### Repeat with only Tumaco

```{r}
t_regression_tests <- c("Age", "Ethnicity", "Sex", "Weight")
t_lm_regression_demographics <- extract_linear_regression(
  t_regression_numeric, query = "Therapeutic_Outcome_Final", factors = t_regression_tests,
  excel = glue("excel/numeric_demographics_regression_tumaco_final_sex_ethnicity_age-v{ver}.xlsx"))
pp(file = "images/demographics_only_tumaco_linear_regression.svg")
t_lm_regression_demographics[["forest"]]
dev.off()
t_lm_regression_demographics[["forest"]]
t_lm_regression_demographics[["summary"]]
```

Now repeat with only Tumaco, this should also be a part of a
supplemental Figure.

```{r}
t_log_regression_demographics <- extract_logistic_regression(
  t_regression_df, query = "Therapeutic_Outcome_Final", factors = t_regression_tests,
  excel = glue("excel/t_multivariable_logistic_regression-v{ver}.xlsx"))
pp(file = glue("figures/t_multivariable_logistic_regression.svg"))
t_log_regression_demographics[["forest"]]
dev.off()
t_log_regression_demographics[["forest"]]
t_log_regression_demographics[["summary"]]

t_log_iterative_regression_demographics <- iterate_logistic_regression(
  t_regression_df, query = "Therapeutic_Outcome_Final", factors = t_regression_tests,
  excel = glue("excel/t_simple_logistic_regression-v{ver}.xlsx"))
pp(file = glue("figures/t_simple_logistic_regression-v{ver}.svg"))
t_log_iterative_regression_demographics[["forest"]]
dev.off()
t_log_iterative_regression_demographics[["forest"]]
t_log_iterative_regression_demographics[["summary"]]
```

If we decide to add things like typeofcells/visitnumber/etc to the
above, then we will absolutely need to use multilevel regression and
use the full combined metadata of the sample sheet and demographics
combined.

```{r}
wanted_queries <- c("Therapeutic_Outcome_Final", "sex", "clinic", "Ethnicity", "Age")
full_meta <- pData(tc_valid)
full_meta_numeric <- full_meta
for (f in wanted_queries) {
  full_meta_numeric[[f]] <- as.numeric(as.factor(full_meta_numeric[[f]]))
}

corheat <- ggstatsplot::ggcorrmat(full_meta_numeric, cor.vars = wanted_queries)
pp(file = "images/type_donor_final_visit_sex_clinic_etnia_age_corheat.pdf")
corheat
dev.off()
corheat
```

### Repeat using a logistical model

In the previous block, the function extract_stepwise_regression() uses
a linear model to extract the f/r values used for the forest plot and
performs a series (or one) stepwise regression.  In the following I
will repeat that process using a logistic regression model.

I bet I wrote it down somewhere above, but wtf is up with the prefix
tdfvscea?

Also, it may be because I have been messing around with wanted_mtrx,
but it should be a set of factors, now numeric.

```{r}
## Also, we are excluding donor
test_factors <- c("typeofcells", "visitnumber", "Sex", "clinic", "Ethnicity", "Age")
logit_regression_test <- extract_logistic_regression(pData(tc_clinical_nobiop), query = "finaloutcome",
                                                     factors = test_factors)

logit_regression_test[["forest"]]
logit_regression_test[["summary"]]
```

The careful reader might notice an odd change in the next block: I
changed from 'Therapeutic_Final_Outcome' to 'finaloutcome'.  This is
because these two data structures have slightly different sources for
the per-person and/or per-sample annotations.  In the first instance
(and all the previous blocks), that information is coming from the
demographics worksheet provided by Maria Adelaida.  In the second
instance (the following block), it is coming from the individual
sample sheet.  The extremely careful reader would also note that there
is a series of blocks in the data structures worksheet which seeks to
ensure that these two data sources agree with each other, and that it
throws a testthat-based hissy-fit if they do not.

```{r}
test_queries <- c("clinic", "Sex", "Ethnicity", "Age", "finaloutcome")
tc_regression_numeric <- pData(tc_clinical_nobiop)[, test_queries]
numeric_mtrx <- pData(t_clinical_nobiop)[, test_queries]
for (f in colnames(numeric_mtrx)) {
  tc_regression_numeric[[f]] <- as.numeric(tc_regression_numeric[[f]])
  numeric_mtrx[[f]] <- as.numeric(numeric_mtrx[[f]])
}

corheat <- ggstatsplot::ggcorrmat(numeric_mtrx, label = TRUE, cor.vars = test_queries)
pp(file = "images/corheat_tumaco_cali.png")
corheat
dev.off()
corheat
```

This correlation heatmap tells us that it is a bad idea to
simultaneously consider Age and Ethnicity in any models we create to
express the data.  It may be possible to make a guess about which is
more appropriate to consider by performing a regression table of each
individually?

Let us try once with all factors included, then remove Ethnicity and
Age sequentially.  It might be wiser to use the quartile'd version of
age?

Fundamentally, this just repeats what we just saw, but makes clearer
that Age and Ethnicity are problematic if placed in the same model.
My reasoning: Theresa suggested that any correlation >= 0.65 is
problematic.

The following block was originally a bit wrong-headed because it was
performed before we realized that we were feeding the lm the full
experimental design with multiple entries per person.

I modified it to use the more appropriate data and decided to leave it
here as a reminder.

```{r}
csea_extracted_regression <- extract_linear_regression(
  tc_regression_numeric, query = "finaloutcome",
  factors = c("clinic", "Sex", "Ethnicity", "Age"), scale = FALSE,
  excel = "excel/tc_regression_table_csea.xlsx")
pp(file = "images/tc_regression_csea.png")
csea_extracted_regression[["forest"]]
dev.off()
csea_extracted_regression[["forest"]]
csea_extracted_regression[["summary"]]

csa_extracted_regression <- extract_linear_regression(
  tc_regression_numeric, query = "finaloutcome",
  factors = c("clinic", "Sex", "Age"),
  excel = "excel/tc_regression_table_csa.xlsx")
pp(file = "images/tc_regression_csa.png")
csa_extracted_regression[["forest"]]
dev.off()
csa_extracted_regression[["forest"]]
csa_extracted_regression[["summary"]]

cse_extracted_regression <- extract_linear_regression(
  tc_regression_numeric, query = "finaloutcome",
  factors = c("clinic", "Sex", "Ethnicity"),
  excel = "excel/tc_regression_table_cse.xlsx")
pp(file = "images/tc_regression_cse.png")
cse_extracted_regression[["forest"]]
dev.off()
cse_extracted_regression[["forest"]]
cse_extracted_regression[["summary"]]
```

## Repeat with only Tumaco

Thus, clinic cannot be in the model.  In addition I will remove
height/weight because we know a priori how they correlate.

```{r}
initial_queries <- c("Sex", "Ethnicity", "Age",
                     "Evolution_Time", "Num_Active_Lesions",
                     "V2_New_Lesions", "V3_New_Lesions",
                     "Adherence", "finaloutcome")
t_initial_mtrx <- pData(t_clinical_nobiop)[, initial_queries]

tumaco_cross <- corr_cross(t_initial_mtrx)
pp(file = "images/tumaco_crosscor.png")
tumaco_cross
dev.off()
tumaco_cross
```

Once again, let us consider a smaller set to identify factors that
should not go into a model test together, we assume that once again
this will prove to be Ethnicity and Age.

```{r}
test_queries <- c("Sex", "Ethnicity", "Age", "finaloutcome")
t_numeric_mtrx <- t_initial_mtrx[, test_queries]
for (f in colnames(t_numeric_mtrx)) {
  t_numeric_mtrx[[f]] <- as.numeric(t_numeric_mtrx[[f]])
}

corheat <- ggstatsplot::ggcorrmat(t_numeric_mtrx, label = TRUE, cor.vars = test_queries)
corheat
```

It turns out they are even more tightly correlated in this subset than in
the full dataset.

## A series of linear regression models

Given the above, I know that I should not include some groups of
factors in a model together, but I want to get a feel for which
factor combinations are most/least informative with respect to final
outcome.

```{r}
t_sea_extracted_regression <- extract_linear_regression(
  t_regression_numeric, query = "Therapeutic_Outcome_Final",
  factors = c("Sex", "Ethnicity", "Age"),
  excel = "excel/t_regression_table_t_sea.xlsx")
pp(file = "images/tc_regression_t_sea.png")
t_sea_extracted_regression[["forest"]]
dev.off()
t_sea_extracted_regression[["forest"]]
t_sea_extracted_regression[["summary"]]

t_sa_extracted_regression <- extract_linear_regression(
  t_regression_numeric, query = "Therapeutic_Outcome_Final",
  factors = c("Sex", "Age"),
  excel = "excel/t_regression_table_t_sa.xlsx")
pp(file = "images/t_regression_sa.png")
t_sa_extracted_regression[["forest"]]
dev.off()
t_sa_extracted_regression[["forest"]]
t_sa_extracted_regression[["summary"]]

t_se_extracted_regression <- extract_linear_regression(
  t_regression_numeric, query = "Therapeutic_Outcome_Final",
  factors = c("Sex", "Ethnicity"),
  excel = "excel/t_regression_table_se.xlsx")
pp(file = "images/t_regression_se.png")
t_se_extracted_regression[["forest"]]
dev.off()
t_se_extracted_regression[["forest"]]
t_se_extracted_regression[["summary"]]
```

# Global views of all cell types

Now that those 'global' metrics are out of the way, lets look at some
global metrics of the data following normalization; the most likely
plots are of course PCA but also a couple of heatmaps.

### Figure 1

Over time the preference for which samples to include in a 'global'
PCA has changed, as well as preferences for how to arrange/label/color
them.  The following shows a couple of perspectives.

```{r}
tc_type <- set_expt_conditions(tc_valid, fact = "typeofcells") %>%
  set_expt_batches(fact = "finaloutcome") %>%
  set_expt_colors(color_choices[["type"]])

tc_norm <- sm(normalize_expt(tc_type, transform = "log2", norm = "quant",
                             convert = "cpm", filter = TRUE))

tc_pca <- plot_pca(tc_norm, plot_labels = FALSE,
                    plot_title = "PCA - Cell type", size_column = "visitnumber")
pp(file = "figures/tc_pca_sized.pdf")
tc_pca
dev.off()
tc_pca
tc_pca <- plot_pca(tc_norm, plot_labels = FALSE,
                   plot_title = "PCA - Cell type")
pp(file = "figures/tc_pca_nosize.pdf")
tc_pca
dev.off()
tc_pca

write.csv(tc_pca[["table"]], file = "excel/tc_donor_pca_coords.csv")
tc_cf_norm <- set_expt_batches(tc_norm, fact = "visitnumber")
tc_cf_corheat <- plot_corheat(tc_cf_norm, plot_title = "Heirarchical clustering:
         cell types")
pp(file = "figures/tc_cf_corheat.svg")
tc_cf_corheat
dev.off()
tc_cf_corheat

tc_cf_disheat <- plot_disheat(tc_cf_norm, plot_title = "Heirarchical clustering:
         cell types")
tc_cf_disheat
```

## Figure 1B: Transcriptomic profiles of primary innate cells

A potential figure legend for the following images might include:

The observed counts per gene for all of the clinical samples were
filtered, log transformed, cpm converted, and quantile normalized.
The colors were defined by cell types and shapes by patient visit.
When the first two principle components were plotted, clustering was
observed by cell type.  The biopsy samples were significantly
different from the innate immune cell types.

```{r}
fig1v2_norm <- normalize_expt(tc_type, transform = "log2",
                              convert = "cpm", norm = "quant", filter = TRUE)
fig1v2_pca <- plot_pca(fig1v2_norm, cis = FALSE)
pp(file = "figures/tc_type_v2.pdf")
fig1v2_pca
dev.off()
fig1v2_pca
```

# Compare samples by clinic

Spoiler alert:  This section will eventually suggest pretty strongly
that we will not easily be able to use the Cali samples.  Thus, after
finishing it, we will likely exclude those samples.

Take a moment to view the biopsy samples.  We separated them by clinic
(Cali or Tumaco), and this view of the samples is the only one which
does not suggest a strong difference between the two clinics.
However, it also suggests that the biopsy samples will not prove very
helpful.

## Biopsies by clinic

There are too few biopsy samples to get a strong view of cure/fail.
In addition, these are 'messier' than any other sample type.  As a
result, it is difficult to discern a pattern in them which help
elucidate cure vs. fail.  If we play with the various parameters used
to perform the count modification via ruv/sva, we get slightly
different views, some more evocative than others; but the following is
our most canonical view.

### Figure 3E: Biopsies

```{r}
tc_biopsies_norm <- normalize_expt(tc_biopsies, transform = "log2",
                                   convert = "cpm", norm = "quant", filter = TRUE)
tc_biopsies_pca <- plot_pca(tc_biopsies_norm)
tc_biopsies_pca
pp(file = "figures/tc_biopsies_pca.svg")
tc_biopsies_pca[["plot"]]
dev.off()

tc_biopsies_nb <- normalize_expt(tc_biopsies, transform = "log2",
                                 convert = "cpm", batch = "svaseq", filter = TRUE)
tc_biopsies_nb_pca <- plot_pca(tc_biopsies_nb)
pp(file = "figures/figure3E_biopsies.svg")
tc_biopsies_nb_pca[["plot"]]
dev.off()
tc_biopsies_nb_pca
```

I worry that we rely too heavily on PCA.

## Patient Race and clinic?

How strong is the effect of ethnicity/ethnicity+clinic?  In the worst
case scenario, these surrogates could make interpreting the results
problematic.  The following blocks will explore that question a little
and I think come to the general conclusion that race and/or clinic are
not significant problems.

### All samples, both clinics

Compared to the cell type effect, clinic/race is, as we already know,
utterly insignificant.  The question still stands, how significant?
There does appear to be an effect in the data which is relevant to
race.  I think if we want to be able to explore this fully, we would
need more people.

```{r}
etnia_expt <- set_expt_conditions(tc_valid, fact = "clinic_etnia") %>%
  set_expt_colors(color_choices[["clinic_etnia"]])
etnia_norm <- normalize_expt(etnia_expt, transform = "log2", convert = "cpm",
                             filter = TRUE, norm = "quant")
plot_pca(etnia_norm)

etnia_nb <- normalize_expt(etnia_expt, transform = "log2", convert = "cpm",
                           filter = TRUE, batch = "svaseq")
plot_pca(etnia_nb)
```

#### Only Tumaco

There is an imbalance in the identity of people who attended each
clinic.  Given that we are focusing on the Tumaco samples, here is the
distribution of race/cell type:

```{r}
t_etnia_norm <- normalize_expt(t_etnia_expt, transform = "log2", convert = "cpm",
                               filter = TRUE, norm = "quant")
plot_pca(t_etnia_norm)

t_etnia_nb <- normalize_expt(t_etnia_expt, transform = "log2", convert = "cpm",
                             filter = TRUE, batch = "svaseq")
plot_pca(t_etnia_nb)
```

### Biopsy samples, both clinics.

The biopsy samples are missing people of indigenous origin who went to
the Tumaco clinic.

```{r}
tc_bp_ec <- set_expt_conditions(tc_biopsies, fact = "clinic_etnia") %>%
  set_expt_colors(color_choices[["clinic_etnia"]])
etnia_bp_norm <- normalize_expt(tc_bp_ec, transform = "log2", convert = "cpm",
                                filter = TRUE, norm = "quant")
plot_pca(etnia_bp_norm)
```

#### Biopsy samples, Tumaco.

The biopsy samples are by far the 'messiest,' that remains true when
considering the ethnicity of the individual patients.

```{r}
t_bp_ec <- set_expt_conditions(tc_biopsies, fact = "etnia") %>%
  set_expt_colors(color_choices[["ethnicity"]])
t_etnia_bp_norm <- normalize_expt(t_bp_ec, transform = "log2", convert = "cpm",
                                  filter = TRUE, norm = "quant")
plot_pca(t_etnia_bp_norm)
```

I think there are not enough samples to try sva with this.

### Eosinophil samples, both clinics.

When we ask the same question of the clinical cell types, it is
possible to see more samples, but not a significantly clearer view of
the race effect on the transcriptional profile.

```{r}
tc_eo_ec <- set_expt_conditions(tc_eosinophils, fact = "clinic_etnia") %>%
  set_expt_colors(color_choices[["clinic_etnia"]])
etnia_eo_norm <- normalize_expt(tc_eo_ec, transform = "log2", convert = "cpm",
                                filter = TRUE, norm = "quant")
plot_pca(etnia_eo_norm)

etnia_eo_nb <- normalize_expt(tc_eo_ec, transform = "log2", convert = "cpm",
                                filter = TRUE, batch = "svaseq")
ethnicity_pca <- plot_pca(etnia_eo_nb)
pp(file = "figures/ethnicity_eo_pca.svg")
ethnicity_pca[["plot"]]
dev.off()
ethnicity_pca[["plot"]]
```

#### Eosinophil samples, Tumaco.

The eosinophils are our least-abundant cell type, as such the view of
ethnicity using them is particularly problematic; but we do at least
have a few samples from each group.  With that in mind, these appear
to show some significant difference among the three groups.

```{r}
t_eo_ec <- set_expt_conditions(t_eosinophils, fact = "etnia") %>%
  set_expt_colors(color_choices[["ethnicity"]])
t_etnia_eo_norm <- normalize_expt(t_eo_ec, transform = "log2", convert = "cpm",
                                filter = TRUE, norm = "quant")
plot_pca(t_etnia_eo_norm)

t_etnia_eo_nb <- normalize_expt(t_eo_ec, transform = "log2", convert = "cpm",
                                filter = TRUE, batch = "svaseq")
plot_pca(t_etnia_eo_nb)
```

### Monocyte samples, both clinics.

In general, the monocytes show the strongest differences in any
comparison we have performed.  This is true in the context of race as
well.  Thus, even before applying sva, we see some separation among
the monocyte samples with respect to ethnicity.

```{r}
tc_mo_ec <- set_expt_conditions(tc_monocytes, fact = "clinic_etnia") %>%
  set_expt_colors(color_choices[["clinic_etnia"]])
etnia_mo_norm <- normalize_expt(tc_mo_ec, transform = "log2", convert = "cpm",
                                filter = TRUE, norm = "quant")
plot_pca(etnia_mo_norm)

etnia_mo_nb <- normalize_expt(tc_mo_ec, transform = "log2", convert = "cpm",
                              filter = TRUE, batch = "svaseq")
etnia_mo_nb_pca <- plot_pca(etnia_mo_nb)
pp(file = "figures/ethnicity_mo_nb_pca.svg")
etnia_mo_nb_pca[["plot"]]
dev.off()
etnia_mo_nb_pca
```

#### Monocyte samples, Tumaco.

The ability to see some separation by ethnicity among the monocyte
samples remains, at least slightly, true when considering only the
Tumaco samples.

```{r}
t_mo_ec <- set_expt_conditions(t_monocytes, fact = "etnia") %>%
  set_expt_colors(color_choices[["ethnicity"]])
t_etnia_mo_norm <- normalize_expt(t_mo_ec, transform = "log2", convert = "cpm",
                                  filter = TRUE, norm = "quant")
plot_pca(t_etnia_mo_norm)

t_etnia_mo_nb <- normalize_expt(t_mo_ec, transform = "log2", convert = "cpm",
                                filter = TRUE, batch = "svaseq")
plot_pca(t_etnia_mo_nb)
```

### Neutrophil samples, both clinics.

In a fashion similar to our other effects, the neutrophils are intermediate.

```{r}
tc_ne_ec <- set_expt_conditions(tc_neutrophils, fact = "clinic_etnia") %>%
    set_expt_colors(color_choices[["clinic_etnia"]])
etnia_ne_norm <- normalize_expt(tc_ne_ec, transform = "log2", convert = "cpm",
                                filter = TRUE, norm = "quant")
plot_pca(etnia_ne_norm)

etnia_ne_nb <- normalize_expt(tc_ne_ec, transform = "log2", convert = "cpm",
                                filter = TRUE, batch = "svaseq")
etnia_ne_nb_pca <- plot_pca(etnia_ne_nb)
pp(file = "figures/ethnicity_ne_nb_pca.svg")
etnia_ne_nb_pca[["plot"]]
dev.off()
etnia_ne_nb_pca
```

#### Neutrophil samples, Tumaco.

The Tumaco-only neutrophils are something of a counter example to the
previous statement.  The easiest to discern race-effect appears to me
to come from the Neutrophils from Tumaco.

```{r}
t_ne_ec <- set_expt_conditions(t_neutrophils, fact = "etnia") %>%
    set_expt_colors(color_choices[["ethnicity"]])
t_etnia_ne_norm <- normalize_expt(t_ne_ec, transform = "log2", convert = "cpm",
                                  filter = TRUE, norm = "quant")
plot_pca(t_etnia_ne_norm)

t_etnia_ne_nb <- normalize_expt(t_ne_ec, transform = "log2", convert = "cpm",
                                filter = TRUE, batch = "svaseq")
plot_pca(t_etnia_ne_nb)
```

## Sex

The imbalances observed with respect to clinic/race are significantly
less profound than those observed with respect to the sex of patients
who participated in the study.  It is almost certainly possible to see
some degree of a sex-based effect in the available transcriptomes.

```{r}
sex_expt <- set_expt_conditions(tc_valid, fact = "sex") %>%
  set_expt_colors(color_choices[["sex"]])
sex_norm <- normalize_expt(sex_expt, transform = "log2", convert = "cpm",
                           filter = TRUE, norm = "quant")
plot_pca(sex_norm)

sex_nb <- normalize_expt(sex_expt, transform = "log2", convert = "cpm",
                         filter = TRUE, batch = "svaseq")
plot_pca(sex_nb)
```

## Sex and clinic

### All samples, both clinics

```{r}
clinic_sex_expt <- set_expt_conditions(tc_valid, fact = "clinic_sex") %>%
  set_expt_colors(color_choices[["clinic_sex"]])
clinic_sex_norm <- normalize_expt(clinic_sex_expt, transform = "log2", convert = "cpm",
                                  filter = TRUE, norm = "quant")
plot_pca(clinic_sex_norm)

clinic_sex_nb <- normalize_expt(clinic_sex_expt, transform = "log2", convert = "cpm",
                             filter = TRUE, batch = "svaseq")
plot_pca(clinic_sex_nb)
```

### Biopsy samples, both clinics.

```{r}
tc_bp_sc <- set_expt_conditions(tc_biopsies, fact = "clinic_sex") %>%
  set_expt_colors(color_choices[["clinic_sex"]])
clinic_sex_bp_norm <- normalize_expt(tc_bp_sc, transform = "log2", convert = "cpm",
                                filter = TRUE, norm = "quant")
plot_pca(clinic_sex_bp_norm)
```

I think there are not enough samples to try sva with this.

### Eosinophil samples, both clinics.

```{r
tc_eo_sc <- set_expt_conditions(tc_eosinophils, fact = "clinic_sex") %>%
  set_expt_colors(color_choices[["clinic_sex"]])
clinic_sex_eo_norm <- normalize_expt(tc_eo_sc, transform = "log2", convert = "cpm",
                                     filter = TRUE, norm = "quant")
plot_pca(clinic_sex_eo_norm)
```

### Monocyte samples, both clinics.

```{r}
tc_mo_clinic_sex <- set_expt_conditions(tc_monocytes, fact = "clinic_sex") %>%
  set_expt_colors(color_choices[["clinic_sex"]])
tc_mo_clinic_sex_norm <- normalize_expt(tc_mo_clinic_sex, transform = "log2", convert = "cpm",
                                        filter = TRUE, norm = "quant")
plot_pca(tc_mo_clinic_sex_norm)

tc_mo_clinic_sex_nb <- normalize_expt(tc_mo_clinic_sex, transform = "log2", convert = "cpm",
                                      filter = TRUE, batch = "svaseq")
plot_pca(tc_mo_clinic_sex_nb)
```

### Neutrophil samples, both clinics.

```{r}
tc_ne_clinic_sex <- set_expt_conditions(tc_neutrophils, fact = "clinic_sex") %>%
    set_expt_colors(color_choices[["clinic_sex"]])
tc_ne_clinic_sex_norm <- normalize_expt(tc_ne_clinic_sex, transform = "log2", convert = "cpm",
                                        filter = TRUE, norm = "quant")
plot_pca(tc_ne_clinic_sex_norm)
```

## Eosinophils by clinic

In contrast, the Eosinophil samples do have significant amounts of
variance which discriminates the two clinics.  At the time of this
writing, there are fewer eosinophil samples than monocytes and
neutrophils; as a result there are no samples which failed from Cali.
This is somewhat limiting is we wish to look for differences between
the cure and fail samples which came from the two clinics.

### Figure 3B: Eosinophils

```{r}
tc_eosinophils_norm <- normalize_expt(tc_eosinophils, transform = "log2",
                                      convert = "cpm", norm = "quant", filter = TRUE)
tc_eosinophils_pca <- plot_pca(tc_eosinophils_norm, plot_labels = FALSE)
tc_eosinophils_pca
pp(file = "figures/tc_eosinophils.svg")
tc_eosinophils_pca[["plot"]]
dev.off()

tc_eosinophils_nb <- normalize_expt(tc_eosinophils, transform = "log2",
                                    convert = "cpm", batch = "svaseq", filter = TRUE)
tc_eosinophils_nb_pca <- plot_pca(tc_eosinophils_nb, plot_labels = FALSE)
pp(file = "figures/figure3B_eosinophils.svg")
tc_eosinophils_nb_pca$plot
dev.off()
tc_eosinophils_nb_pca$plot
```

## Monocytes by clinic

In contrast with the eosinophil samples, we have one patient's
monocyte and neutrophil samples which did not cure.  As we will see,
there is one person from Cali who did not cure, this person is not
different with respect to tracscriptome than the other people from Cali.

### Figure 3C: Monocytes

```{r}
tc_monocytes_norm <- normalize_expt(tc_monocytes, transform = "log2",
                                       convert = "cpm", norm = "quant", filter = TRUE)
tc_monocytes_pca <- plot_pca(tc_monocytes_norm, plot_labels = FALSE)
tc_monocytes_pca
pp(file = "figures/tc_monocytes_pca.svg")
tc_monocytes_pca[["plot"]]
dev.off()

tc_monocytes_nb <- normalize_expt(tc_monocytes, transform = "log2",
                                  convert = "cpm", batch = "svaseq", filter = TRUE)
tc_monocytes_nb_pca <- plot_pca(tc_monocytes_nb, plot_labels = FALSE)
pp(file = "figures/figure3C_monocytes.svg")
tc_monocytes_nb_pca$plot
dev.off()
tc_monocytes_nb_pca$plot
```

## Neutrophils by clinic

Finally, that same one person does appear to be different than the
others from Cali when looking at neutrophils.

### Figure 3D: Neutrophils

```{r}
tc_neutrophils_norm <- normalize_expt(tc_neutrophils, transform = "log2",
                                      convert = "cpm", norm = "quant", filter = TRUE)
tc_neutrophils_pca <- plot_pca(tc_neutrophils_norm, plot_labels = FALSE)
tc_neutrophils_pca
pp(file = "figures/tc_neutrophils_pca.svg")
tc_neutrophils_pca[["plot"]]
dev.off()

tc_neutrophils_nb <- normalize_expt(tc_neutrophils, transform = "log2",
                                    convert = "cpm", batch = "svaseq", filter = TRUE)
tc_neutrophils_nb_pca <- plot_pca(tc_neutrophils_nb, plot_labels = FALSE)
pp(file = "figures/figure3D_neutrophils.svg")
tc_neutrophils_nb_pca$plot
dev.off()
tc_neutrophils_nb_pca$plot
```

## PCA: Compare clinics

Now that we have these various subsets, perform an explicit comparison
of the samples which came from the two clinics.

### Figure 3, panel A: 'ALL Samples'

```{r}
tc_clinic_type <- tc_valid %>%
  set_expt_conditions(fact = "clinic") %>%
  set_expt_batches(fact = "typeofcells")

tc_clinic_type_norm <- normalize_expt(tc_clinic_type, transform = "log2", convert = "cpm",
                                      norm = "quant", filter = TRUE)
tc_clinic_type_pca <- plot_pca(tc_clinic_type_norm)
tc_clinic_type_pca
tc_clinic_type_nb <- normalize_expt(tc_clinic_type, transform = "log2", convert = "cpm",
                                    batch = "svaseq", filter = TRUE)
tc_clinic_type_nb_pca <- plot_pca(tc_clinic_type_nb)
tc_clinic_type_nb_pca
pp(file = "figures/figure3a_all_samples.svg")
tc_clinic_type_nb_pca$plot
dev.off()
```


```{r}
tc_clinical_norm <- sm(normalize_expt(tc_clinical, filter = "simple", transform = "log2",
                                      norm = "quant", convert = "cpm"))
clinical_pca <- plot_pca(tc_clinical_norm, plot_labels = FALSE,
                         cis = NULL,
                         plot_title = "PCA - clinical samples")
clinical_pca

tc_clinical_nb <- normalize_expt(tc_clinical, filter = "simple", transform = "log2",
                                 batch = "svaseq", convert = "cpm")
tc_clinical_nb_pca <- plot_pca(tc_clinical_nb)
tc_clinical_nb_pca$plot

clinical_pca_info <- pca_information(
    tc_clinical_norm, plot_pcas = TRUE, num_components = 30,
    expt_factors = c("visitnumber", "typeofcells", "finaloutcome",
                   "clinic", "donor"))
clinical_pca_info$anova_neglogp_heatmap
clinical_pca_info$pca_plots$PC4_PC7

clinical_scores <- pca_highscores(tc_clinical_norm)
clinical_scores[["highest"]][,"Comp.4"]

test_factors <- c("visitnumber", "typeofcells", "finaloutcome",
                  "clinic", "sex", "etnia")
clinical_varpart <- simple_varpart(tc_clinical, factors = test_factors)
clinical_varpart
```

## Iterative SVA followed by PCA

Another way to explore the effect of SVA is to iteratively increase
the number of SVs removed by it and look at some simple plots of the
resulting data.  Ideally, this should complement the comparison of
individual SVs vs. PCs performed by Theresa (spoiler alert, I think it
did).

```{r}
first <- normalize_expt(tc_clinical, transform = "log2", convert = "cpm",
                        filter = TRUE, batch = "svaseq", surrogates = 1)
first_info <- pca_information(
    first, plot_pcas = TRUE, num_components = 30,
    expt_factors = c("visitnumber", "typeofcells",
                     "finaloutcome", "clinic"))
first_info$anova_neglogp_heatmap
first_info$pca_plots[["PC1_PC2"]]

second <- normalize_expt(tc_clinical, transform = "log2", convert = "cpm",
                         filter = TRUE, batch = "svaseq", surrogates = 2) %>%
  set_expt_batches(fact = "clinic")
second_info <- pca_information(
    second, plot_pcas = TRUE, num_components = 30,
    expt_factors = c("visitnumber", "typeofcells",
                     "finaloutcome", "clinic"))
second_info$anova_neglogp_heatmap

third <- normalize_expt(tc_clinical, transform = "log2", convert = "cpm",
                        filter = TRUE, batch = "svaseq", surrogates = 3) %>%
  set_expt_batches(fact = "clinic")
third_info <- pca_information(
    third, plot_pcas = TRUE, num_components = 30,
    expt_factors = c("visitnumber", "typeofcells",
                     "finaloutcome", "clinic"))
third_info$anova_neglogp_heatmap

fourth <- normalize_expt(tc_clinical, transform = "log2", convert = "cpm",
                         filter = TRUE, batch = "svaseq", surrogates = 4) %>%
  set_expt_batches(fact = "clinic")
fourth_info <- pca_information(
    fourth, plot_pcas = TRUE, num_components = 30,
    expt_factors = c("visitnumber", "typeofcells",
                     "finaloutcome", "clinic"))
fourth_info$anova_neglogp_heatmap
fourth_info[["pca_plots"]][["PC1_PC2"]]

fifth <- normalize_expt(tc_clinical, transform = "log2", convert = "cpm",
                        filter = TRUE, batch = "svaseq", surrogates = 5) %>%
  set_expt_batches(fact = "clinic")
fifth_info <- pca_information(
    fifth, plot_pcas = TRUE, num_components = 30,
    expt_factors = c("visitnumber", "typeofcells",
                     "finaloutcome", "clinic"))
fifth_info$anova_neglogp_heatmap
fifth_info[["pca_plots"]][["PC1_PC12"]]

sixth <- normalize_expt(tc_clinical, transform = "log2", convert = "cpm",
                        filter = TRUE, batch="svaseq", surrogates = 6) %>%
  set_expt_batches(fact = "clinic")
sixth_info <- pca_information(
    sixth, plot_pcas = TRUE, num_components = 30,
    expt_factors = c("visitnumber", "typeofcells",
                     "finaloutcome", "clinic"))
sixth_info$anova_neglogp_heatmap

seventh <- normalize_expt(tc_clinical, transform = "log2", convert = "cpm",
                          filter = TRUE, batch = "svaseq", surrogates = 7) %>%
  set_expt_batches(fact = "clinic")
seventh_info <- pca_information(
    seventh, plot_pcas = TRUE, num_components = 30,
    expt_factors = c("visitnumber", "typeofcells",
                     "finaloutcome", "clinic"))
seventh_info$anova_neglogp_heatmap

eighth <- normalize_expt(tc_clinical, transform = "log2", convert = "cpm",
                        filter = TRUE, batch = "svaseq", surrogates = 8)
eighth_info <- pca_information(
    eighth, plot_pcas = TRUE, num_components = 30,
    expt_factors = c("visitnumber", "typeofcells",
                     "finaloutcome", "clinic"))
eighth_info$anova_neglogp_heatmap
```

# Variance Partition

variancePartition (@hoffmanVariancePartitionInterpretingDrivers2016)
provides a nice toolbox of methods to examine the relationship between
various metadata factors in a dataset with respect to the variance
observed in the dataset's expression.  We usually use it as a quick
way to see the relative likelihood that a differential expression
of various factors will provide useful/helpful output.

```{r}
## Mostly running twice to make sure that reordering the factors does not affect the end result.
tc_varpart <- simple_varpart(
  tc_clinical_nobiop, factors = c("visitnumber", "typeofcells",
                                  "finaloutcome", "clinic", "sex", "etnia"))
tc_varpart

tc_varpartv2 <- simple_varpart(
  tc_clinical_nobiop, factors = c("donor", "visitnumber", "typeofcells",
                                  "finaloutcome"))
pp(file = "images/tc_donor_visit_type_finaloutcome_varpart.pdf")
tc_varpartv2
dev.off()
tc_varpartv2

tc_varpartv3 <- simple_varpart(
  tc_clinical_nobiop, factors = c("donor", "visitnumber", "typeofcells"))
pp(file = "images/tc_donor_visit_type_varpart.pdf")
tc_varpartv3
dev.off()
tc_varpartv3

tc_varpartv4 <- simple_varpart(
  tc_clinical_nobiop, factors = c("finaloutcome", "sex", "Ethnicity"))
pp(file = "images/tc_final_sex_ethnicity_varpart.pdf")
tc_varpartv4
dev.off()
tc_varpartv4

t_varpartv3 <- simple_varpart(
  t_clinical_nobiop, factors = c("donor", "visitnumber", "typeofcells"))
pp(file = "images/t_donor_visit_type_varpart.pdf")
t_varpartv3
dev.off()
t_varpartv3

c_varpartv3 <- simple_varpart(
  c_clinical, factors = c("donor", "visitnumber", "typeofcells"))
pp(file = "images/c_donor_visit_type_varpart.pdf")
c_varpartv3
dev.off()
c_varpartv3
```

## Some factors of later interest

Maria Adelaida asked about using variancePartition to query a few
other factors in the Cali, Tumaco, and both datasets.  These factors
include: Sex, Age, Ethnicity, Clinic; and potentially Adherence, time
of evolution, and previous diagnosis.

I am not sure if those factors are already in the expressionset
metadata, but if not we can certainly bring them back.  In the
following block I will therefore repeat a simple variancePartition
analysis using first the full dataset (Tumaco+Cali), then each clinic
alone; in each instance I will do one round with sex, ethnicity, age,
and clinic followed by the same and finaloutcome (as a reference point
to something we are already looking at).

```{r}
table(pData(tc_clinical_nobiop)$typeofcells)
table(pData(t_clinical_nobiop)$typeofcells)
table(pData(c_clinical_nobiop)$typeofcells)

tc_fun_varpart <- simple_varpart(tc_clinical_nobiop,
                                 factors = c("sex", "etnia", "Age", "clinic"))
pp(file = "images/tc_fun_varpart.pdf")
tc_fun_varpart
dev.off()
tc_fun_varpart
tc_fun_outcome_varpart <- simple_varpart(tc_clinical_nobiop,
                                         factors = c("sex", "etnia", "Age", "clinic", "finaloutcome"))
pp(file = "images/tc_fun_outcome_varpart.pdf")
tc_fun_outcome_varpart
dev.off()
tc_fun_outcome_varpart
c_fun_varpart <- simple_varpart(c_clinical_nobiop,
                                factors = c("sex", "etnia", "Age"))
pp(file = "images/c_fun_varpart.pdf")
c_fun_varpart
dev.off()
c_fun_outcome_varpart <- simple_varpart(c_clinical_nobiop,
                                        factors = c("sex", "etnia", "Age", "finaloutcome"))
pp(file = "images/c_fun_outcome_varpart.pdf")
c_fun_outcome_varpart
dev.off()
c_fun_outcome_varpart
t_fun_varpart <- simple_varpart(t_clinical_nobiop,
                                factors = c("sex", "etnia", "Age"))
pp(file = "images/t_fun_varpart.pdf")
t_fun_varpart
dev.off()
t_fun_varpart
t_fun_outcome_varpart <- simple_varpart(t_clinical_nobiop,
                                        factors = c("sex", "etnia", "Age", "finaloutcome"))
pp(file = "images/t_fun_outcome_varpart.pdf")
t_fun_outcome_varpart
dev.off()
t_fun_outcome_varpart
```

## Visualize: Repeat plots using only the Tumaco samples

The following should be a nearly copy/pasted version of the above, but
limited to the Tumaco samples.

### All samples

#### Figure xx panels A+B

```{r}
t_clinical_nobiop_norm <- normalize_expt(t_clinical_nobiop, filter = TRUE, norm = "quant",
                                         convert = "cpm", transform = "log2")
t_clinical_nobiop_pca <- plot_pca(t_clinical_nobiop_norm, plot_labels = FALSE)
pp(file = "figures/t_clinical_nobiop_pca.svg")
t_clinical_nobiop_pca[["plot"]]
dev.off()
t_clinical_nobiop_pca

t_clinical_nobiop_nb <- normalize_expt(t_clinical_nobiop, filter = TRUE, convert = "cpm",
                                       transform = "log2", batch = "svaseq")
t_clinical_nobiop_nb_pca <- plot_pca(t_clinical_nobiop_nb, plot_labels = FALSE)
pp(file = "figures/t_clinical_nobiop_sva_pca.svg")
t_clinical_nobiop_nb_pca[["plot"]]
dev.off()
t_clinical_nobiop_nb_pca
```

#### Figure xx panels C+D

```{r}
tc_clinical_nobiop_norm <- normalize_expt(tc_clinical_nobiop, filter = TRUE, norm = "quant",
                                          convert = "cpm", transform = "log2")
tc_clinical_nobiop_pca <- plot_pca(tc_clinical_nobiop_norm, plot_labels = FALSE)
pp(file = "figures/tc_clinical_nobiop_pca.svg")
tc_clinical_nobiop_pca[["plot"]]
dev.off()
tc_clinical_nobiop_pca

tc_clinical_nobiop_nb <- normalize_expt(tc_clinical_nobiop, filter = TRUE, convert = "cpm",
                                        transform = "log2", batch = "svaseq")
tc_clinical_nobiop_nb_pca <- plot_pca(tc_clinical_nobiop_nb, plot_labels = FALSE)
pp(file = "figures/tc_clinical_nobiop_sva_pca.svg")
tc_clinical_nobiop_nb_pca[["plot"]]
dev.off()
tc_clinical_nobiop_nb_pca
```

Now we have a new, smaller set of primary samples which are
categorized by cell type.

### Visualize: Biopsy samples only Tumaco

The biopsy samples remain basically impenetrable. I think it
would be particularly nice if we could judge cure/fail from a visit 1
biopsy.

```{r}
t_biopsies_norm <- normalize_expt(t_biopsies, transform = "log2", convert = "cpm",
  norm = "quant", filter = TRUE)

t_biopsies_pca <- plot_pca(t_biopsies_norm,
  plot_labels = FALSE)
t_biopsies_pca

t_biopsies_nb <- normalize_expt(t_biopsies, transform = "log2", convert = "cpm",
                                batch = "svaseq", filter = TRUE)
t_biopsies_nb_pca <- plot_pca(t_biopsies_nb, plot_labels = FALSE)
t_biopsies_nb_pca$plot
```

### Visualize: Monocyte samples only Tumaco

In contrast, I suspect that we can get meaningful data from the
other cell types.  The monocyte samples are still a bit messy.

### Figure 4A: Monocytes

```{r}
t_monocyte_norm <- normalize_expt(t_monocytes, transform = "log2", convert = "cpm",
                                  norm = "quant", filter = TRUE)

t_monocyte_pca <- plot_pca(t_monocyte_norm,
  plot_labels = FALSE)
t_monocyte_pca

t_monocyte_nb <- normalize_expt(t_monocytes, transform = "log2", convert = "cpm",
                                batch = "svaseq", filter = TRUE)
t_monocyte_nb_pca <- plot_pca(t_monocyte_nb, plot_labels = FALSE)
pp(file = "figures/figure4A_monocytes.svg")
t_monocyte_nb_pca$plot
dev.off()
t_monocyte_nb_pca$plot
```

### Visualize: Neutrophil samples only Tumaco

### Figure 4A: Neutrophils

```{r}
t_neutrophil_norm <- normalize_expt(t_neutrophils, transform = "log2", convert = "cpm",
                                    norm = "quant", filter = TRUE)

t_neutrophil_pca <- plot_pca(t_neutrophil_norm,
                             plot_labels = FALSE)
t_neutrophil_pca

t_neutrophil_nb <- normalize_expt(t_neutrophils, transform = "log2", convert = "cpm",
                                     batch = "svaseq", filter = TRUE)
t_neutrophil_nb_pca <- plot_pca(t_neutrophil_nb, plot_labels = FALSE)
pp(file = "figures/figure4A_neutrophils.svg")
t_neutrophil_nb_pca$plot
dev.off()
t_neutrophil_nb_pca$plot
```

### Visualize: Eosinophil samples only Tumaco

### Figure 4A: Eosinophils

```{r}
t_eosinophil_norm <- normalize_expt(t_eosinophils, transform = "log2", convert = "cpm",
                                    norm = "quant", filter = TRUE)

t_eosinophil_pca <- plot_pca(t_eosinophil_norm,
                             plot_labels = FALSE)
t_eosinophil_pca

t_eosinophil_nb <- normalize_expt(t_eosinophils, transform = "log2", convert = "cpm",
                                  batch = "svaseq", filter = TRUE)
t_eosinophil_nb_pca <- plot_pca(t_eosinophil_nb, plot_labels = FALSE)
pp(file = "figures/figure4A_eosinophils.svg")
t_eosinophil_nb_pca$plot
dev.off()
t_eosinophil_nb_pca$plot
```

### Visualize: Look at Cell types C/F by visit

#### Monocytes, Visit 1

```{r}
t_monocyte_v1 <- subset_expt(t_monocytes, subset = "visitnumber=='1'")
t_monocyte_v1_norm <- normalize_expt(t_monocyte_v1, norm = "quant", convert = "cpm",
                                     transform = "log2", filter = TRUE)
t_monocyte_v1_pca <- plot_pca(t_monocyte_v1_norm, plot_labels = FALSE)
t_monocyte_v1_pca

t_monocyte_v1_nb <- normalize_expt(t_monocyte_v1, convert = "cpm",
                                   transform = "log2", filter = TRUE, batch = "svaseq")
t_monocyte_v1_nb_pca <- plot_pca(t_monocyte_v1_nb, plot_labels = FALSE)
t_monocyte_v1_nb_pca$plot
```

#### Monocytes Visit 2

```{r}
t_monocyte_v2 <- subset_expt(t_monocytes, subset = "visitnumber=='2'")
t_monocyte_v2_norm <- normalize_expt(t_monocyte_v2, norm = "quant", convert = "cpm",
                                     transform = "log2", filter = TRUE)
t_monocyte_v2_pca <- plot_pca(t_monocyte_v2_norm, plot_labels = FALSE)
t_monocyte_v2_pca$plot

t_monocyte_v2_nb <- normalize_expt(t_monocyte_v2, convert = "cpm",
                                   transform = "log2", filter = TRUE, batch = "svaseq")
t_monocyte_v2_nb_pca <- plot_pca(t_monocyte_v2_nb, plot_labels = FALSE)
t_monocyte_v2_nb_pca$plot
```

#### Monocytes Visit 3

```{r}
t_monocyte_v3 <- subset_expt(t_monocytes, subset = "visitnumber=='3'")
t_monocyte_v3_norm <- normalize_expt(t_monocyte_v3, norm = "quant", convert = "cpm",
                                   transform = "log2", filter = TRUE)
t_monocyte_v3_pca <- plot_pca(t_monocyte_v3_norm, plot_labels = FALSE)
t_monocyte_v3_pca

t_monocyte_v3_nb <- normalize_expt(t_monocyte_v3, convert = "cpm",
                                   transform = "log2", filter = TRUE, batch = "svaseq")
t_monocyte_v3_nb_pca <- plot_pca(t_monocyte_v3_nb, plot_labels = FALSE)
t_monocyte_v3_nb_pca$plot
```

#### Neutrophils, Visit 1

```{r} neutrophils_by_visit_v1}
t_neutrophil_v1 <- subset_expt(t_neutrophils, subset = "visitnumber=='1'")
t_neutrophil_v1_norm <- normalize_expt(t_neutrophil_v1, norm = "quant", convert = "cpm",
                                   transform = "log2", filter = TRUE)
t_neutrophil_v1_pca <- plot_pca(t_neutrophil_v1_norm, plot_labels = FALSE)
t_neutrophil_v1_pca$plot

t_neutrophil_v1_nb <- normalize_expt(t_neutrophil_v1, convert = "cpm",
                                     transform = "log2", filter = TRUE, batch = "ruvg")
t_neutrophil_v1_nb_pca <- plot_pca(t_neutrophil_v1_nb, plot_labels = FALSE)
t_neutrophil_v1_nb_pca$plot
```

#### Neutrophils Visit 2

```{r}
t_neutrophil_v2 <- subset_expt(t_neutrophils, subset = "visitnumber=='2'")
t_neutrophil_v2_norm <- normalize_expt(t_neutrophil_v2, norm = "quant", convert = "cpm",
                                       transform = "log2", filter = TRUE)
t_neutrophil_v2_pca <- plot_pca(t_neutrophil_v2_norm, plot_labels = FALSE)
t_neutrophil_v2_pca

t_neutrophil_v2_nb <- normalize_expt(t_neutrophil_v2, convert = "cpm",
                                     transform = "log2", filter = TRUE, batch = "svaseq")
t_neutrophil_v2_nb_pca <- plot_pca(t_neutrophil_v2_nb, plot_labels = FALSE)
t_neutrophil_v2_nb_pca$plot
```

#### Neutrophils Visit 3

```{r}
t_neutrophil_v3 <- subset_expt(t_neutrophils, subset = "visitnumber=='3'")
t_neutrophil_v3_norm <- normalize_expt(t_neutrophil_v3, norm = "quant", convert = "cpm",
                                       transform = "log3", filter = TRUE)
t_neutrophil_v3_pca <- plot_pca(t_neutrophil_v3_norm, plot_labels = FALSE)
t_neutrophil_v3_pca

t_neutrophil_v3_nb <- normalize_expt(t_neutrophil_v3, convert = "cpm",
                                     transform = "log2", filter = TRUE, batch = "svaseq")
t_neutrophil_v3_nb_pca <- plot_pca(t_neutrophil_v3_nb, plot_labels = FALSE)
t_neutrophil_v3_nb_pca$plot
```

#### Eosinophils, Visit 1

```{r}
t_eosinophil_v1 <- subset_expt(t_eosinophils, subset = "visitnumber=='1'")
t_eosinophil_v1_norm <- normalize_expt(t_eosinophil_v1, norm = "quant", convert = "cpm",
                                   transform = "log2", filter = TRUE)
t_eosinophil_v1_pca <- plot_pca(t_eosinophil_v1_norm, plot_labels = FALSE)
t_eosinophil_v1_pca

t_eosinophil_v1_nb <- normalize_expt(t_eosinophil_v1, convert = "cpm",
                                     transform = "log2", filter = TRUE, batch = "svaseq")
t_eosinophil_v1_nb_pca <- plot_pca(t_eosinophil_v1_nb, plot_labels = FALSE)
t_eosinophil_v1_nb_pca$plot
```

#### Eosinophils Visit 2

```{r}
t_eosinophil_v2 <- subset_expt(t_eosinophils, subset = "visitnumber=='2'")
t_eosinophil_v2_norm <- normalize_expt(t_eosinophil_v2, norm = "quant", convert = "cpm",
                                       transform = "log2", filter = TRUE)
t_eosinophil_v2_pca <- plot_pca(t_eosinophil_v2_norm, plot_labels = FALSE)
t_eosinophil_v2_pca

t_eosinophil_v2_nb <- normalize_expt(t_eosinophil_v2, convert = "cpm",
                                     transform = "log2", filter = TRUE, batch = "svaseq")
t_eosinophil_v2_nb_pca <- plot_pca(t_eosinophil_v2_nb, plot_labels = FALSE)
t_eosinophil_v2_nb_pca$plot
```

#### Eosinophils Visit 3

```{r}
t_eosinophil_v3 <- subset_expt(t_eosinophils, subset = "visitnumber=='3'")
t_eosinophil_v3_norm <- normalize_expt(t_eosinophil_v3, norm = "quant", convert = "cpm",
                                       transform = "log3", filter = TRUE)
t_eosinophil_v3_pca <- plot_pca(t_eosinophil_v3_norm, plot_labels = FALSE)
t_eosinophil_v3_pca

t_eosinophil_v3_nb <- normalize_expt(t_eosinophil_v3, convert = "cpm",
                                     transform = "log2", filter = TRUE, batch = "svaseq")
t_eosinophil_v3_nb_pca <- plot_pca(t_eosinophil_v3_nb, plot_labels = FALSE)
t_eosinophil_v3_nb_pca$plot
```

## Recategorize: Concatenate cure/fail and cell type

In the following block the experimental condition was reset to the
concatenation of clinical outcome and type of cells.  There are an
insufficient number of biopsy samples for them to be useful in this
visualization, so they are ignored.

```{r}
desired_levels <- c("cure_biopsy", "failure_biopsy", "cure_eosinophils", "failure_eosinophils",
                    "cure_monocytes", "failure_monocytes", "cure_neutrophils", "failure_neutrophils")
new_fact <- factor(
    paste0(pData(t_clinical)[["condition"]], "_",
           pData(t_clinical)[["batch"]]),
    levels = desired_levels)

t_clinical_concat <- set_expt_conditions(t_clinical, fact = new_fact) %>%
  set_expt_batches(fact = "visitnumber") %>%
  set_expt_colors(color_choices[["cf_type"]]) %>%
  subset_expt(subset="typeofcells!='biopsy'")

## Try to ensure that the levels stay in the order I want
meta <- pData(t_clinical_concat) %>%
  mutate(condition = fct_relevel(condition, desired_levels))
pData(t_clinical_concat) <- meta
```

### Visualize: Look at Tumaco-only samples by cell type and cure/fail

The following block is pretty wild to my eyes; it seems to me that the
variances introduced by cell type basically wipe out the apparent
differences between cure/fail that we were able to see previously.

I suppose this is not entirely surprising, but when we had the Cali
samples it at least looked like there were differences which were
explicitly between cure/fail across cell types.  I suppose this means
those differences were actually coming from the unbalanced state of
the two clinics from the perspective of clinic.

```{r}
t_clinical_concat_norm <- normalize_expt(t_clinical_concat, transform = "log2", convert = "cpm",
                                       norm = "quant", filter = TRUE)
t_clinical_concat_norm_pca <- plot_pca(t_clinical_concat_norm)
t_clinical_concat_norm_pca$plot

t_clinical_concat_nb <- normalize_expt(t_clinical_concat, transform = "log2", convert = "cpm",
                                       batch = "svaseq", filter = TRUE)
t_clinical_concat_nb_pca <- plot_pca(t_clinical_concat_nb)
t_clinical_concat_nb_pca$plot
```

# Visit comparisons

Let us shift the focus from cell type and/or Cure/Fail to the visit
number.  As you are likely aware, the three visits are significantly
spread apart according to the clinical treatment of each patient.
Thus we will now separate the samples by visit in order to more easily
see what new patterns emerge.

## Recategorize: All visits together

Now let us shift the view slightly to focus on changes observed over time.

I have a note from Maria Adelaida that she would like to flesh this
section out with some more pdf versions of various pre/post SVA plots.
If I understood/wrote down correctly her goals:

1.  3 visits, all cell types.
2.  3 visits, all clinical cell types (e.g. no biopsies), only Tumaco
3.  #1, #2 after sva
4.  Repeat the only C/F by visit with/out SVA and make pretty versions.

```{r}
tc_visit_expt <- set_expt_conditions(tc_clinical, fact = "visitnumber") %>%
  set_expt_batches(fact = "finaloutcome") %>%
  set_expt_colors(color_choices[["visit2"]])
tc_visit_norm <- normalize_expt(tc_visit_expt, filter = TRUE, transform = "log2",
                                convert = "cpm", norm = "quant")
tc_visit_norm_pca <- plot_pca(tc_visit_norm)
pp(file = "images/tc_visit_norm_alltypes.pdf")
tc_visit_norm_pca$plot
dev.off()
tc_visit_norm_pca

tc_visit_nb <- normalize_expt(tc_visit_expt, filter = TRUE, transform = "log2",
                              convert = "cpm", batch = "svaseq")
tc_visit_nb_pca <- plot_pca(tc_visit_nb)
pp(file = "images/tc_visit_sva_alltypes.pdf")
tc_visit_nb_pca$plot
dev.off()
tc_visit_nb_pca

##  Repeat for only Tumaco
t_visit_expt <- subset_expt(tc_clinical, subset = "clinic=='tumaco'") %>%
  set_expt_conditions(fact = "visitnumber") %>%
  set_expt_batches(fact = "finaloutcome") %>%
  set_expt_colors(color_choices[["visit2"]])
t_visit_norm <- normalize_expt(t_visit_expt, filter = TRUE, transform = "log2",
                                convert = "cpm", norm = "quant")
t_visit_norm_pca <- plot_pca(t_visit_norm)
pp(file = "images/t_visit_norm_alltypes.pdf")
t_visit_norm_pca$plot
dev.off()
t_visit_norm_pca

t_visit_nb <- normalize_expt(t_visit_expt, filter = TRUE, transform = "log2",
                             convert = "cpm", batch = "svaseq")
t_visit_nb_pca <- plot_pca(t_visit_nb)
pp(file = "images/t_visit_sva_alltypes.pdf")
t_visit_nb_pca$plot
dev.off()
t_visit_nb_pca

## Finally, limit to only the clinical celltypes
t_visit_clinical_expt <- subset_expt(t_visit_expt, subset = "typeofcells!='biopsy'")
t_visit_clinical_norm <- normalize_expt(t_visit_clinical_expt, filter = TRUE, transform = "log2",
                                        convert = "cpm", norm = "quant")
t_visit_clinical_norm_pca <- plot_pca(t_visit_clinical_norm)
pp(file = "images/t_visit_clinical_norm_alltypes.pdf")
t_visit_clinical_norm_pca$plot
dev.off()
t_visit_clinical_norm_pca

t_visit_clinical_nb <- normalize_expt(t_visit_clinical_expt, filter = TRUE,
                                      transform = "log2", convert = "cpm", batch = "svaseq")
t_visit_clinical_nb_pca <- plot_pca(t_visit_clinical_nb)
pp(file = "images/t_visit_nobiop_sva_alltypes.pdf")
t_visit_clinical_nb_pca$plot
dev.off()
t_visit_clinical_nb_pca
```

When looking at all cell types, it is quite difficult to see
differences among the three visits.

## Visualize: C/F for only the visit 1 samples

Wen we had both Cali and Tumaco samples, it looked like there was
variance suggesting differences between cure and fail for visit 1.  I
think the following block will suggest pretty strongly that this was
not true.

```{r}
tv1_samples <- set_expt_batches(tv1_samples, fact = "typeofcells")
tv1_norm <- normalize_expt(tv1_samples, transform = "log2", convert = "cpm",
                          norm = "quant", filter = TRUE)
tv1_pca <- plot_pca(tv1_norm)
pp(file = "images/tv1_pca.pdf")
tv1_pca
dev.off()
tv1_pca$plot

tv1_nb <- normalize_expt(tv1_samples, transform = "log2", convert = "cpm",
                         filter = TRUE, batch = "svaseq")
tv1_nb_pca <- plot_pca(tv1_nb, plot_labels = FALSE)
pp(file = "images/tv1_sva_pca.pdf")
tv1_nb_pca$plot
dev.off()
tv1_nb_pca
```

## Visualize: C/F for only the visit 2 samples

```{r}
tv2_samples <- set_expt_batches(tv2_samples, fact = "typeofcells")
tv2_norm <- normalize_expt(tv2_samples, transform = "log2", convert = "cpm",
                          norm = "quant", filter = TRUE)
tv2_pca <- plot_pca(tv2_norm)
pp(file = "images/tv2_pca.pdf")
tv2_pca
dev.off()
tv2_pca$plot

tv2_nb <- normalize_expt(tv2_samples, transform = "log2", convert = "cpm",
                         filter = TRUE, batch = "svaseq")
tv2_nb_pca <- plot_pca(tv2_nb, plot_labels = FALSE)
pp(file = "images/tv2_sva_pca.pdf")
tv2_nb_pca$plot
dev.off()
tv2_nb_pca
```

## Visualize: C/F for only the visit 3 samples

```{r}
tv3_samples <- set_expt_batches(tv3_samples, fact = "typeofcells")
tv3_norm <- normalize_expt(tv3_samples, transform = "log2", convert = "cpm",
                          norm = "quant", filter = TRUE)
tv3_pca <- plot_pca(tv3_norm)
pp(file = "images/tv3_pca.pdf")
tv3_pca
dev.off()
tv3_pca$plot

tv3_nb <- normalize_expt(tv3_samples, transform = "log2", convert = "cpm",
                         filter = TRUE, batch = "svaseq")
tv3_nb_pca <- plot_pca(tv3_nb, plot_labels = FALSE)
pp(file = "images/tv3_sva_pca.pdf")
tv3_nb_pca$plot
dev.off()
tv3_nb_pca
```

### Visualize: Comparing 3 visits by cell type

Separate the samples by cell type in order to more easily observe
patterns with respect to visit and clinical outcome.

#### Monocytes across visits

In the following few blocks we are coloring the samples by visit and
final outcome.  We are also separating the three primary celltypes of
interest.  If I understand correctly, Maria Adelaida has an interest
in a nice version of each of these 6 plots (normalized pca
before/after SVA for each celltype).

```{r}
t_visitcf_monocyte_norm <- normalize_expt(t_visitcf_monocyte, norm = "quant", convert = "cpm",
                                transform = "log2", filter = TRUE)
t_visitcf_monocyte_pca <- plot_pca(t_visitcf_monocyte_norm, plot_labels = FALSE)
pp(file = "images/t_monocyte_visitcf_norm_pca.pdf")
t_visitcf_monocyte_pca$plot
dev.off()
t_visitcf_monocyte_pca

t_visitcf_monocyte_disheat <- plot_disheat(t_visitcf_monocyte_norm)
t_visitcf_monocyte_disheat$plot

t_visitcf_monocyte_nb <- normalize_expt(t_visitcf_monocyte, convert = "cpm",
                                    transform = "log2", filter = TRUE, batch = "svaseq")
t_visitcf_monocyte_nb_pca <- plot_pca(t_visitcf_monocyte_nb, plot_labels = FALSE)
pp(file = "images/t_monocyte_visitcf_sva_pca.pdf")
t_visitcf_monocyte_nb_pca$plot
dev.off()
t_visitcf_monocyte_nb_pca
```

#### Eosinophils across visits

Repeat the above with Eosinophils, we should therefore have slightly
fewer glyphs on the plot.

```{r}
t_visitcf_eosinophil_norm <- normalize_expt(t_visitcf_eosinophil, norm = "quant", convert = "cpm",
                                transform = "log2", filter = TRUE)
t_visitcf_eosinophil_pca <- plot_pca(t_visitcf_eosinophil_norm, plot_labels = FALSE)
pp(file = "images/t_eosinophil_visitcf_norm_pca.pdf")
t_visitcf_eosinophil_pca$plot
dev.off()
t_visitcf_eosinophil_pca

t_visitcf_eosinophil_disheat <- plot_disheat(t_visitcf_eosinophil_norm)
t_visitcf_eosinophil_disheat$plot

t_visitcf_eosinophil_nb <- normalize_expt(t_visitcf_eosinophil, convert = "cpm",
                                    transform = "log2", filter = TRUE, batch = "svaseq")
t_visitcf_eosinophil_nb_pca <- plot_pca(t_visitcf_eosinophil_nb, plot_labels = FALSE)
pp(file = "images/t_eosinophil_visitcf_sva_pca.pdf")
t_visitcf_eosinophil_nb_pca$plot
dev.off()
t_visitcf_eosinophil_nb_pca
```

#### Neutrophils across visits

```{r}
t_visitcf_neutrophil_norm <- normalize_expt(t_visitcf_neutrophil, norm = "quant", convert = "cpm",
                                transform = "log2", filter = TRUE)
t_visitcf_neutrophil_pca <- plot_pca(t_visitcf_neutrophil_norm, plot_labels = FALSE)
pp(file = "images/t_neutrophil_visitcf_norm_pca.pdf")
t_visitcf_neutrophil_pca$plot
dev.off()
t_visitcf_neutrophil_pca

t_visitcf_neutrophil_disheat <- plot_disheat(t_visitcf_neutrophil_norm)
t_visitcf_neutrophil_disheat$plot

t_visitcf_neutrophil_nb <- normalize_expt(t_visitcf_neutrophil, convert = "cpm",
                                          transform = "log2", filter = TRUE, batch = "svaseq")
t_visitcf_neutrophil_nb_pca <- plot_pca(t_visitcf_neutrophil_nb, plot_labels = FALSE)
pp(file = "images/t_neutrophil_visitcf_sva_pca.pdf")
t_visitcf_neutrophil_nb_pca$plot
dev.off()
t_visitcf_neutrophil_nb_pca
```

#### Celltypes by visit, C/F batch: Monocytes

We are backing off the granular view of visit and Fail/Cure in the
following block and instead just considering the three visits.  This
previously only considered the normalized result, now we wish to add
the sva modified result and print out pdfs thereof.  Once again, we
are repeating 3 times, once for each cell type.

```{r}
t_visit_monocyte <- set_expt_conditions(t_visitcf_monocyte, prefix = "v",
                                        fact = "visitnumber") %>%
  set_expt_batches("finaloutcome") %>%
  set_expt_colors(color_choices[["visit"]])
t_visit_monocyte_norm <- normalize_expt(t_visit_monocyte,
                                        transform = "log2", convert = "cpm",
                                        norm = "quant", filter = TRUE)

t_visit_monocyte_norm_pca <- plot_pca(t_visit_monocyte_norm, plot_labels = FALSE)
pp(file = "figures/t_monocyte_visit_norm_pca.svg")
t_visit_monocyte_norm_pca[["plot"]]
dev.off()
t_visit_monocyte_norm_pca
t_visitcf_monocyte_nb <- normalize_expt(t_visitcf_monocyte,
                                        transform = "log2", convert = "cpm",
                                        batch = "svaseq", filter = TRUE)
t_visitcf_monocyte_nb_pca <- plot_pca(t_visitcf_monocyte_nb, plot_labels = FALSE)
pp(file = "figures/t_monocyte_visit_sva_pca.svg")
t_visitcf_monocyte_nb_pca[["plot"]]
dev.off()
t_visitcf_monocyte_nb_pca
```

#### Celltypes by visit, C/F batch: Eosinophils

```{r}
t_visit_eosinophil <- set_expt_conditions(t_visitcf_eosinophil, prefix = "v",
                                            fact = "visitnumber") %>%
  set_expt_batches("finaloutcome") %>%
  set_expt_colors(color_choices[["visit"]])
t_visit_eosinophil_norm <- normalize_expt(t_visit_eosinophil,
                                          transform = "log2", convert = "cpm",
                                          norm = "quant", filter = TRUE)
t_visit_eosinophil_norm_pca <- plot_pca(t_visit_eosinophil_norm, plot_labels = FALSE)
pp(file = "figures/t_eosinophil_visit_norm_pca.svg")
t_visit_eosinophil_norm_pca[["plot"]]
dev.off()
t_visit_eosinophil_norm_pca
t_visit_eosinophil_nb <- normalize_expt(t_visit_eosinophil,
                                        transform = "log2", convert = "cpm",
                                        batch = "svaseq", filter = TRUE)
t_visit_eosinophil_nb_pca <- plot_pca(t_visit_eosinophil_nb, plot_labels = FALSE)
pp(file = "figures/t_eosinophil_visit_sva_pca.svg")
t_visit_eosinophil_nb_pca[["plot"]]
dev.off()
t_visit_eosinophil_nb_pca
```

#### Celltypes by visit, C/F batch: Neutrophils

```{r}
t_visit_neutrophil <- set_expt_conditions(t_visitcf_neutrophil, prefix = "v",
                                          fact = "visitnumber") %>%
  set_expt_batches("finaloutcome") %>%
  set_expt_colors(color_choices[["visit"]])
t_visit_neutrophil_norm <- normalize_expt(t_visit_neutrophil,
                                          transform = "log2", convert = "cpm",
                                          norm = "quant", filter = TRUE)
t_visit_neutrophil_norm_pca <- plot_pca(t_visit_neutrophil_norm, plot_labels = FALSE)
pp(file = "figures/t_neutrophil_visit_norm_pca.svg")
t_visit_neutrophil_norm_pca[["plot"]]
dev.off()
t_visit_neutrophil_norm_pca
t_visit_neutrophil_nb <- normalize_expt(t_visit_neutrophil,
                                        transform = "log2", convert = "cpm",
                                        batch = "svaseq", filter = TRUE)
t_visit_neutrophil_nb_pca <- plot_pca(t_visit_neutrophil_nb, plot_labels = FALSE)
pp(file = "figures/t_neutrophil_visit_sva_pca.svg")
t_visit_neutrophil_nb_pca[["plot"]]
dev.off()
t_visit_neutrophil_nb_pca
```

# Persistence

### Take a look

See if there are any patterns which look usable.

```{r, eval=FALSE}
## All
t_persistence_norm <- normalize_expt(t_persistence, transform = "log2", convert = "cpm",
                                   norm = "quant", filter = TRUE)
plot_pca(t_persistence_norm)$plot
t_persistence_nb <- normalize_expt(t_persistence, transform = "log2", convert = "cpm",
                                 batch = "svaseq", filter = TRUE)
plot_pca(t_persistence_nb)$plot

## Biopsies
##persistence_biopsy_norm <- normalize_expt(persistence_biopsy, transform = "log2", convert = "cpm",
##                                   norm = "quant", filter = TRUE)
##plot_pca(persistence_biopsy_norm)$plot
## Insufficient data

## Monocytes
t_persistence_monocyte_norm <- normalize_expt(t_persistence_monocyte, transform = "log2", convert = "cpm",
                                              norm = "quant", filter = TRUE)
plot_pca(t_persistence_monocyte_norm)$plot
t_persistence_monocyte_nb <- normalize_expt(t_persistence_monocyte, transform = "log2", convert = "cpm",
                                 batch = "svaseq", filter = TRUE)
plot_pca(t_persistence_monocyte_nb)$plot

## Neutrophils
t_persistence_neutrophil_norm <- normalize_expt(t_persistence_neutrophil, transform = "log2", convert = "cpm",
                                                norm = "quant", filter = TRUE)
plot_pca(t_persistence_neutrophil_norm)$plot
t_persistence_neutrophil_nb <- normalize_expt(t_persistence_neutrophil, transform = "log2", convert = "cpm",
                                 batch = "svaseq", filter = TRUE)
plot_pca(t_persistence_neutrophil_nb)$plot

## Eosinophils
t_persistence_eosinophil_norm <- normalize_expt(t_persistence_eosinophil, transform = "log2", convert = "cpm",
                                   norm = "quant", filter = TRUE)
plot_pca(t_persistence_eosinophil_norm)$plot
t_persistence_eosinophil_nb <- normalize_expt(t_persistence_eosinophil, transform = "log2", convert = "cpm",
                                 batch = "svaseq", filter = TRUE)
plot_pca(t_persistence_eosinophil_nb)$plot
```

# Classify me!

I wrote out all the z2.2 and z2.3 specific variants to a couple files,
I want to see if I can classify a human sample as infected with 2.2 or
2.3.

```{r, eval=FALSE}
z22 <- read.csv("csv/variants_22.csv")
z23 <- read.csv("csv/variants_23.csv")
cure <- read.csv("csv/cure_variants.txt")
fail <- read.csv("csv/fail_variants.txt")
z22_vec <- gsub(pattern="\\-", replacement="_", x=z22[["x"]])
z23_vec <- gsub(pattern="\\-", replacement="_", x=z23[["x"]])
cure_vec <- gsub(pattern="\\-", replacement="_", x=cure)
fail_vec <- gsub(pattern="\\-", replacement="_", x=fail)

classify_zymo <- function(sample) {
  arbitrary_tags <- sm(readr::read_tsv(sample))
  arbitrary_ids <- arbitrary_tags[["position"]]
  message("Length: ", length(arbitrary_ids), ", z22: ",
          sum(arbitrary_ids %in% z22_vec) / (length(z22_vec)), " z23: ",
          sum(arbitrary_ids %in% z23_vec) / (length(z23_vec)))
}

arbitrary_sample <- "preprocessing/TMRC30156/outputs/40freebayes_lpanamensis_v36/all_tags.txt.xz"
classify_zymo(arbitrary_sample)
```

# Visualizing composite scores

First lets get the gene IDs and colors for these plots.

```{r}
library(viridis)
wanted_genes <- c("IFI44L", "IFI27", "PRR5", "PRR5-ARHGAP8", "RHCE",
                  "FBXO39", "RSAD2", "SMTNL1", "USP18", "AFAP1")
wanted_idx <- fData(tc_valid)[["hgnc_symbol"]] %in% wanted_genes
wanted_ids <- rownames(fData(tc_valid))[wanted_idx]
```

## All samples, all visits

```{r}
few <-  subset_genes(tc_valid, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp

few <-  subset_genes(t_clinical, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp
```

## All samples, visit 1

```{r}
few <- subset_genes(tc_clinical, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  subset_expt(subset = "visitnumber=='1'") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp

few <- subset_genes(t_clinical, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  subset_expt(subset = "visitnumber=='1'") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp
```

## Eosinophils, all times

```{r}
few <-  subset_genes(tc_eosinophils, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp

few <-  subset_genes(t_eosinophils, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp
```

## Eosinophils, v1

```{r}
few <-  subset_genes(tc_eosinophils, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  subset_expt(subset = "visitnumber=='1'") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp

few <-  subset_genes(t_eosinophils, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  subset_expt(subset = "visitnumber=='1'") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp
```

## Monocytes all

```{r}
few <-  subset_genes(tc_monocytes, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp

few <-  subset_genes(t_monocytes, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp
```

## Monocytes v1

```{r}
few <-  subset_genes(tc_monocytes, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  subset_expt(subset = "visitnumber=='1'") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp

few <-  subset_genes(t_monocytes, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  subset_expt(subset = "visitnumber=='1'") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp
```

## Neutrophils all

```{r}
few <-  subset_genes(tc_neutrophils, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp

few <-  subset_genes(t_neutrophils, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp
```

## Neutrophils v1

```{r}
few <-  subset_genes(tc_neutrophils, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  subset_expt(subset = "visitnumber=='1'") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp

few <-  subset_genes(t_neutrophils, ids = wanted_ids, method = "keep") %>%
  set_expt_conditions(fact = "finaloutcome") %>%
  subset_expt(subset = "visitnumber=='1'") %>%
  normalize_expt(transform = "log2", convert = "rpkm",
                 column = "mean_cds_len")
shp <- plot_sample_heatmap(few, heatmap_colors = viridis, row_label = wanted_genes)
shp
```

# An external dataset

Let us look at a moderately similar Biopsy dataset of braziliensis
infected individuals.  First, lets do a quick plot of their data, our
biopsies, then combine them.

## Load the data

```{r}
## Load the scott-only and the scott+tmrc3 data
load(glue("rda/tmrc3_external_cf-v{ver}.rda"))
load(glue("rda/tmrc3_external-v{ver}.rda"))
```

## Visualize the two datasets individually

```{r}
our_biopsies <- set_expt_conditions(t_biopsies, "finaloutcome") %>%
  set_expt_colors(color_choices[["cf"]])
our_biopsies_norm <- normalize_expt(our_biopsies, filter = TRUE, transform = "log2",
                                    convert = "cpm", batch = "svaseq")
plot_pca(our_biopsies_norm)$plot

scott_biopsies_norm <- normalize_expt(external_cf, filter = TRUE, transform = "log2",
                                      convert = "cpm", batch = "svaseq")
plot_pca(scott_biopsies_norm)$plot
```

## Visualize them together

```{r}
both_norm <- normalize_expt(tmrc3_external, filter = TRUE, transform = "log2",
                             convert = "cpm", norm = "quant")
plot_pca(both_norm)$plot
both_nb <- normalize_expt(tmrc3_external, filter = TRUE, transform = "log2",
                           convert = "cpm", batch = "svaseq")
plot_pca(both_nb)$plot

external_species <- set_expt_conditions(tmrc3_external, fact = "ParasiteSpecies") %>%
  subset_expt(subset = "ParasiteSpecies!='notapplicable'") %>%
  set_expt_batches(fact = "lab")

tt <- normalize_expt(external_species, transform = "log2", convert = "cpm", norm = "quant",
                     filter = TRUE)
plot_pca(tt)
ttt <- normalize_expt(external_species, transform = "log2", convert = "cpm", batch = "svaseq",
                     filter = TRUE)
plot_pca(ttt)
```

# Parasite distribution

I am resurrecting some of the comparisons of the parasite transcriptome in the host data.

```{r}
lp_cf <- set_expt_conditions(lp_expt, fact = "finaloutcome")

table(pData(lp_cf)[["typeofcells"]])

lp_cf_norm <- normalize_expt(lp_cf, transform = "log2", convert = "cpm",
                             norm = "quant", filter = TRUE)
lp_cf_sm <- plot_sm(lp_cf_norm)
lp_cf_sm
lp_cf_corheat <- plot_corheat(lp_cf_norm)
lp_cf_corheat
lp_cf_norm_pca <- plot_pca(lp_cf_norm)
lp_cf_norm_pca
pp(file = "figures/lp_cf_norm_pca.svg")
lp_cf_norm_pca[["plot"]]
dev.off()

lp_cf_nb <- normalize_expt(lp_cf, transform = "log2", convert = "cpm",
                           batch = "svaseq", filter = "simple")
lp_cf_nb_pca <- plot_pca(lp_cf_nb)
lp_cf_nb_pca
```

Note, the previous task includes visits 2/3 and multiple cell types
and as a result is likely to include the most profoundly infected
people (only those in whom we observe >30,000 reads and >3,000 genes
of parasite reads.  Thus, even though it sort of looks like there
_might_ be a C/F difference, the sva shows that to be a lie.

Nonetheless, we can make this clearer by excluding the visits2/3
and/or non-biopsies.

```{r}
lp_cf_biop <- subset_expt(lp_cf, subset = "typeofcells=='Biopsy'")

lp_cf_biop_norm <- normalize_expt(lp_cf_biop, transform = "log2", convert = "cpm",
                                  norm = "quant", filter = TRUE)
lp_cf_biop_sm <- plot_sm(lp_cf_biop_norm)
lp_cf_biop_sm
lp_cf_biop_corheat <- plot_corheat(lp_cf_biop_norm)
lp_cf_biop_corheat
lp_cf_biop_norm_pca <- plot_pca(lp_cf_biop_norm)
lp_cf_biop_norm_pca

lp_cf_biop_nb <- normalize_expt(lp_cf_biop, transform = "log2", convert = "cpm",
                                batch = "svaseq", filter = "simple")
lp_cf_biop_nb_pca <- plot_pca(lp_cf_biop_nb)
lp_cf_biop_nb_pca
```

# Examining PCs and SVs vs. the metadata: SV loadings

This is coming out of the 09varcor_regression document and was
initially performed by Theresa.  If this works well and is sufficient,
I might remove that document and therefore have that much less stuff
to check on for correctness.

<span style="color:#006400">
Note from atb: I need to make a few changes to this section, primarily
we need to be able to automatically generate the tables of
f-statistics; in case the data changes (which it did since Theresa
performed this, one sample was removed I think).  With that caveat,
the following is coming directly out of her SVA_V3_Tumaco document.  I
also would like to compare the SV-fstats to similar metrics I took of
PCs vs. metadata factors.  My assumption (if I understand the math in
sva at all) is that they should largely complement/agree with each
other.
</span>

We would like to know what the heck SVA is actually correcting for
when we do an SVA correction. Are there any metadatas that these SVs
are correlated with?

To do this, I will run SVA to get the SV loadings. I will then do
something akin to PC loadings analysis to see how these individual SVs
(and combinatorial SVs) are associated with any

I will use a computed F-statistic for this association to measure the
between:within cluster variance in a model (and tell us if that factor
is a good indicator of separation based on that sv loading).

$$\begin{equation}
F-statistic = \frac{TSS - RSS}{RSS}
\end{equation}$$

So for this, I will use a series of linear regressions which model
each dimension of SVA as a function of the observed variables that
describe the known underlying group structure (clinic, visit, patient,
...)

$$\begin{equation}
\underbrace{X_i}_\text{dimension i of SVA} = \underbrace{B_0 + B_1
celltype/visit/clinic/donor}_\text{underlying group structure}
\end{equation}$$

We can do this breakdown in a few ways to answer different questions
which I will explore further below.

We have decided the Cali samples don't offer a lot of extra
information for us, and there is significant clinic batch effect, so
we are going to remove the Cali samples and evaluate the SV loadings.

The first thing to do is the actual SVA to get the loadings.

<span style="color:#006400">
I may have changed a few of Theresa's variable names when I first
copy/pasted this document together without taking note of the
modification; but I am reasonably certain that the intended data
structures are the same.
</span>

```{r}
clinic_sva <- normalize_expt(t_clinical, filter = TRUE)
pheno <- pData(clinic_sva)
edata <- exprs(clinic_sva)

mod <- model.matrix(~as.factor(finaloutcome), data = pheno)
mod0 <- model.matrix(~1, data = pheno)

svobj <- sva::svaseq(edata, mod, mod0)
```

### SV 1

SVA found 4 SVs. We can plot them individually to visually inspect
their separation w.r.t some metadata.

```{r}
svs <- as.data.frame(svobj$sv)
colnames(svs) <- paste0("sv_", seq(1:4))
svs <- cbind(svs, pheno)

sv1_typeofcells <- ggplot(svs, aes(y = sv_1, x = typeofcells, fill = typeofcells)) +
  geom_violin() +
  geom_point(alpha = 0.75) +
  xlab("Type of Cells") +
  ylab("SV 1")  +
  theme_classic() +
  theme(legend.position = "none")

sv1_visit <-  ggplot(svs, aes(y = sv_1, x = visitnumber, fill = visitnumber)) +
  geom_violin() +
  geom_point(alpha = 0.75) +
  xlab("Visit Number") +
  ylab("SV 1")  +
  theme_classic() +
  theme(legend.position = "none")

sv1_donor <- ggplot(svs, aes(y = sv_1, x = donor, fill = donor)) +
  geom_violin() +
  geom_point(alpha = 0.75) +
  xlab("Donor") +
  ylab("SV 1")  +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

sv1_typeofcells
sv1_visit
sv1_donor
##grid.arrange(sv1_typeofcells, sv1_visit, sv1_donor, nrow = 2)
```

### SV2

```{r}
sv2_typeofcells <- ggplot(svs, aes(y = sv_2, x = typeofcells, fill = typeofcells)) +
  geom_violin() +
  geom_point(alpha = 0.75) +
  xlab("Type of Cells") +
  ylab("SV 2")  +
  theme_classic() +
  theme(legend.position = "none")

sv2_visit <-  ggplot(svs, aes(y = sv_2, x = visitnumber, fill = visitnumber)) +
  geom_violin() +
  geom_point(alpha = 0.75) +
  xlab("Visit Number") +
  ylab("SV 2")  +
  theme_classic() +
  theme(legend.position = "none")

sv2_donor <- ggplot(svs, aes(y = sv_2, x = donor, fill = donor)) +
  geom_violin() +
  geom_point(alpha = 0.75) +
  xlab("Donor") +
  ylab("SV 2")  +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))


#grid.arrange(sv2_typeofcells, sv2_visit, sv2_donor, nrow = 2)
sv2_typeofcells
sv2_visit
sv2_donor
```

<span style="color:#006400">
I spent a little time to simplify and try to make the reasoning above
a little more robust so that I can regenerate Theresa's xlsx table of
f-statistics as well as add a little more information.  The following
block attempts this...
</span>

Najib correctly pointed out that I left off clinic in this first invocation.

```{r}
queries <- c("typeofcells", "visitnumber", "clinic", "donor")
tc_clinical_fpstats <- svpc_fstats(tc_clinical, num_pcs = 5, queries = queries)
queries <- c("typeofcells", "visitnumber", "donor")
t_clinical_fpstats <- svpc_fstats(t_clinical, num_pcs = 5, queries = queries)
c_clinical_fpstats <- svpc_fstats(c_clinical, num_pcs = 5, queries = queries)
```

# Send to an xlsx workbook

<span style="color:#006400">
I am going to add a little code in this section to send this to an
xlsx file.  I might need to add a little bit of code as well because I
am not certain that there is a document which contains this
calculation for each data subset.

I put together a quick function which writes the results of one of
these analyses to a xlsx file, but it very much assumes a single
dataset and is not easily amendable to multiple; therefore I will
strip the code out here into a new function to repeat itself for the
Tumaco/Cali/Both data for an arbitrary combination.
</span>

Query from Maria Adelaida: Perform a similar f/p statistics plot/xlsx
table but using the first 5 PCs and SVs; perhaps also include the
amount of variance remaining tale (I forget its name: residuals).

But also do slightly different plots: 2 plots: 1 with PCs before SVA
followed by the SVs, the 1 with SVs followed by post PCs.

Given this, perform this task with: Clinic, Donor, Visit, Celltype
using the clinical samples (no biopsies).

```{r}
write_combined_fpstats <- function(both = tc_clinical_fpstats, tumaco = t_clinical_fpstats,
                                   cali = c_clinical_fpstats,
                                   excel = "excel/combined_svpc_fstats.xlsx") {
  xlsx <- init_xlsx(excel)
  wb <- xlsx[["wb"]]
  excel_basename <- xlsx[["basename"]]
  do_excel <- TRUE
  if (is.null(wb)) {
    do_excel <- FALSE
  }

  current_row <- 1

  pref <- both[["pre_f"]]
  svf <- both[["sv_f"]]
  postf <- both[["post_f"]]
  ## Changing the rownames due to rbind rownames shenanigans.
  rownames(pref) <- paste0("PrePC", seq_len(nrow(pref)))
  rownames(postf) <- paste0("PostPC", seq_len(nrow(postf)))
  allf <- rbind(pref, svf, postf)

  prep <- both[["pre_p"]]
  svp <- both[["sv_p"]]
  postp <- both[["post_p"]]
  rownames(prep) <- paste0("PrePC", seq_len(nrow(prep)))
  rownames(postp) <- paste0("PostPC", seq_len(nrow(postp)))
  allp <- rbind(prep, svp, postp)

  fun_plot <- heatmap.3(as.matrix(allp), dendrogram = "none",
                        scale = "none", trace = "none",
                        Colv = FALSE, Rowv = FALSE)
  image <- grDevices::recordPlot()

  xlsx_result <- write_xlsx(data = allf, wb = wb, sheet = "Fvalues", start_row = current_row,
                            title = "Both clinics, SVA and PC analysis, F-values")
  xlsx_result <- write_xlsx(data = allp, wb = wb, sheet = "Pvalues", start_row = current_row,
                            title = "Both clinics, SVA and PC analysis, P-values")
  current_row <- xlsx_result[["end_row"]] + 2
  try_result <- xlsx_insert_png(
    a_plot = image, wb = wb, sheet = "Pvalues", start_col = ncol(allp) + 2)
  image_files = c()
  if (! "try-error" %in% class(try_result)) {
    image_files = try_result[["filename"]]
  }

  pref <- tumaco[["pre_f"]]
  svf <- tumaco[["sv_f"]]
  postf <- tumaco[["post_f"]]
  ## Changing the rownames due to rbind rownames shenanigans.
  rownames(pref) <- paste0("PrePC", seq_len(nrow(pref)))
  rownames(postf) <- paste0("PostPC", seq_len(nrow(postf)))
  allf <- rbind(pref, svf, postf)

  prep <- tumaco[["pre_p"]]
  svp <- tumaco[["sv_p"]]
  postp <- tumaco[["post_p"]]
  rownames(prep) <- paste0("PrePC", seq_len(nrow(prep)))
  rownames(postp) <- paste0("PostPC", seq_len(nrow(postp)))
  allp <- rbind(prep, svp, postp)

  xlsx_result <- write_xlsx(data = allf, wb = wb, sheet = "Fvalues", start_row = current_row,
                            title = "Tumaco, SVA and PC analysis, F-values")
  xlsx_result <- write_xlsx(data = allp, wb = wb, sheet = "Pvalues", start_row = current_row,
                            title = "Tumaco, SVA and PC analysis, P-values")
  current_row <- xlsx_result[["end_row"]] + 2

  pref <- cali[["pre_f"]]
  svf <- cali[["sv_f"]]
  postf <- cali[["post_f"]]
  ## Changing the rownames due to rbind rownames shenanigans.
  rownames(pref) <- paste0("PrePC", seq_len(nrow(pref)))
  rownames(postf) <- paste0("PostPC", seq_len(nrow(postf)))
  allf <- rbind(pref, svf, postf)

  prep <- cali[["pre_p"]]
  svp <- cali[["sv_p"]]
  postp <- cali[["post_p"]]
  rownames(prep) <- paste0("PrePC", seq_len(nrow(prep)))
  rownames(postp) <- paste0("PostPC", seq_len(nrow(postp)))
  allp <- rbind(prep, svp, postp)

  xlsx_result <- write_xlsx(data = allf, wb = wb, sheet = "Fvalues", start_row = current_row,
                            title = "Cali, SVA and PC analysis, F-values")
  xlsx_result <- write_xlsx(data = allp, sheet = "Pvalues", wb = wb, start_row = current_row,
                            title = "Cali, SVA and PC analysis, P-values")
  current_row <- xlsx_result[["end_row"]] + 2

  excel_ret <- try(openxlsx::saveWorkbook(wb, excel, overwrite = TRUE))
  removed <- try(suppressWarnings(file.remove(image_files)), silent = TRUE)
}

clinical_fpstats <- write_combined_fpstats(
  both = tc_clinical_fpstats, tumaco = t_clinical_fpstats, cali = c_clinical_fpstats,
  excel = glue("excel/clinical_fpstats-v{ver}.xlsx"))
```

The F-stat resulting from an anova for the model sv ~ metadata_factor
shows that the main thing we are correcting for with an SVA correction
(with cure/fail as the model factor) is the cell type. The factor
donor contributes the next highest separation, with clinic falling in
third. the visit contributes essentially no variance in this data,
which we knew from the DE results.

# Bibliography
