---
title: "TMRC3 `r Sys.getenv('VERSION')`: Exploring WGCNA"
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
runtime: shiny
output:
  html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    keep_md: false
    mode: selfcontained
    number_sections: true
    self_contained: true
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

<style type="text/css">
body .main-container {
  max-width: 1600px;
}
body, td {
  font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
  font-size: 16px
}
</style>

```{r options, include=FALSE}
library(hpgltools)
library(dplyr)
library(forcats)
library(glue)
library(tidyverse)

knitr::opts_knit$set(progress = TRUE, verbose = TRUE, width = 90, echo = TRUE)
knitr::opts_chunk$set(
  error = TRUE, fig.width = 8, fig.height = 8, fig.retina = 2,
  out.width = "100%", dev = "png",
  dev.args = list(png = list(type = "cairo-png")))
old_options <- options(digits = 4, stringsAsFactors = FALSE, knitr.duplicate.label = "allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size = 12))
ver <- Sys.getenv("VERSION")
previous_file <- ""
rundate <- format(Sys.Date(), format = "%Y%m%d")

##tmp <- try(sm(loadme(filename = gsub(pattern = "\\.Rmd", replace = "\\.rda\\.xz", x = previous_file))))
rmd_file <- "05wgcna.Rmd"
savefile <- gsub(pattern = "\\.Rmd", replace = "\\.rda\\.xz", x = rmd_file)
loaded <- load(file = glue("rda/tmrc3_data_structures-v{ver}.rda"))
```

# Working to get an initial understanding of WGCNA

I am using Alejandro's document to get a feeling for how WGCNA is
getting modules relevant to cure/fail.

I am reasonably certain he is in turn using this document as his
input:

https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0

# Setting up

```{r}
threads <- WGCNA::enableWGCNAThreads(nThreads = 8)
```

# Input data

The input used is the cell-type specific, tmaco-only, rpkm, modified
by sva.

In the following block I am breaking down what Alejandro did into
smaller pieces so that I can make certain I understand what happened.

```{r}
input <- normalize_expt(t_monocytes, filter = TRUE, batch = "svaseq") %>%
  normalize_expt(convert = "rpkm", column = "mean_cds_len", na_to_zero = TRUE)

wgcna_input <- as.data.frame(exprs(input))
wgcna_input[["ENSEMBLID"]] <- rownames(wgcna_input)

wgcna_melted <- wgcna_input %>%
  gather(key = "samples", value = "counts", -ENSEMBLID)

wgcna_with_meta <- wgcna_melted %>%
  inner_join(., pData(input), by = c("samples" = "tmrcidentifier"))

wgcna_selected <- wgcna_with_meta %>%
  select("ENSEMBLID", "samples", "counts") %>%
  spread(key = "samples", value = "counts")
```

Unless I am mistaken, the above just converted the matrix of counts
into a merged/melted copy of same with the metadata, then removed the
metadata and returned us back to the original state? hmmm...


```{r}
good_samples_genes <- WGCNA::goodSamplesGenes(t(exprs(input)))
summary(good_samples_genes)
good_samples_genes[["allOK"]]

l2input <- normalize_expt(input, transform = "log2")

power_test <- c(c(1:10), seq(from = 12, to = 20, by = 2))
threshold_search <- WGCNA::pickSoftThreshold(
  t(exprs(l2input)), powerVector = power_test,
  networkType = "signed", verbose = 5)

a1 <- ggplot(threshold_search[["fitIndices"]], aes(Power, SFT.R.sq, label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0.8, color = 'red') +
  labs(x = 'Power', y = 'Scale free topology model fit, signed R^2')
a1

a2 <- ggplot(threshold_search[["fitIndices"]], aes(Power, mean.k., label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  labs(x = 'Power', y = 'Mean Connectivity')
a2
```

# Create Modules

```{r}
chosen_power <- 8
## WGCNA calls cor() without specifying its own namespace, so overwrite cor for the moment.
cor <- WGCNA::cor
initial_modules <- WGCNA::blockwiseModules(
  t(exprs(l2input)), maxBlockSize = 11000, TOMType = "signed",
  power = chosen_power, mergeCutHeight = 0.25, numericLabels = FALSE,
  verbose = 3)
cor <- stats::cor

initial_eigen <- initial_modules[["MEs"]]
```

# View initial modules

```{r}
network_colors <- WGCNA::labels2colors(initial_modules[["colors"]])
WGCNA::plotDendroAndColors(
  initial_modules[["dendrograms"]][[1]],
  network_colors[initial_modules[["blockGenes"]][[1]]],
  "Modules",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05)

WGCNA::plotDendroAndColors(
  initial_modules[["dendrograms"]][[1]],
  cbind(initial_modules[["unmergedColors"]], initial_modules[["colors"]]),
  c("unmerged", "merged"),
  dendroLabels = FALSE,
  addGuide = TRUE,
  hang = 0.03,
  guideHang = 0.05)

WGCNA::plotDendroAndColors(
  initial_modules[["dendrograms"]][[1]],
  initial_modules[["colors"]],
  "ME",
  dendroLabels = FALSE,
  addGuide = TRUE,
  hang= 0.03,
  cex.colorLabels = 2,
  marAll = c(1, 5, 3, 1),
  main = ("WGCNA Cluster Dendrogram"),
  guideHang = 0.05)
```

# Consensus reordering

This does not appear to work, FIXME

```{r, eval=FALSE}
initial_reorder <- WGCNA::consensusOrderMEs(
  initial_eigen, useAbs = FALSE,
  useSets = NULL, greyLast = TRUE,
  greyName = paste(WGCNA::moduleColor.getMEprefix(), "grey", sep = ""),
  method = "consensus")
```

# Cross reference against metadata

```{r}
meta_numeric <- data.frame(
  "cf_numeric" = as.numeric(as.factor(pData(l2input)[["finaloutcome"]])),
  "visit_numeric" = as.numeric(as.factor(pData(l2input)[["visitnumber"]])))
rownames(meta_numeric) <- rownames(pData(l2input))

meta_factors <- pData(l2input)[, c("finaloutcome", "visitnumber")]
meta_eigen <- merge(initial_eigen, meta_factors, by = "row.names")
rownames(meta_eigen) <- meta_eigen[["Row.names"]]
meta_eigen[["Row.names"]] <- NULL
kappa <- irr::kappam.fleiss(meta_eigen)

module_trait_corr <- stats::cor(initial_eigen, meta_numeric, use = "p")
module_trait_corr
module_trait_pvalues <- WGCNA::corPvalueStudent(module_trait_corr, nrow(meta_numeric))
module_trait_pvalues
```

# Plot the 'correlations'

On my computer at least, there seems to be difficulty installing the
CorLvelPlot library, so I will just remove this piece of Alejandro's
code for now.

```{r}
module_trait_merged <- merge(initial_eigen, meta_numeric, by = "row.names")
rownames(module_trait_merged) <- module_trait_merged[["Row.names"]]
module_trait_merged[["Row.names"]] <- NULL

#CorLevelPlot::CorLevelPlot(
#  module_trait_merged,
#  x = names(module_trait_merged)[1:18],
#  rotLabX = 90,
#  y = names(module_trait_merged)[19:20],
#  posColKey = "top",
#  col = c("blue1", "skyblue", "white", "pink", "red"))
```

# Extract genes from 'significant' modules

It appears that the modules 'turqoise' and 'pink' are likely the most
interesting.  We can extract the genes from them:

```{r}
wanted <- initial_modules[["colors"]] == "turqoise" | initial_modules[["colors"]] == "pink"
sum(wanted)
interesting_genes <- names(initial_modules[["colors"]])[wanted]

fData(l2input)[interesting_genes, "hgnc_symbol"]
written_interesting <- write_xlsx(fData(l2input)[interesting_genes, ],
                                  excel = glue("excel/wgcna_interesting_genes-v{ver}.xlsx"))

## Note that we can do similarity matrices on the samples too in order to get
## dendrograms which may get interesting groups of samples?
not_grey <- initial_modules[["colors"]] != "grey"
not_grey_exprs <- t(exprs(l2input))[, not_grey]
dim(not_grey_exprs)
not_grey_genes <- colnames(not_grey_exprs)
dist_tom <- 1 - WGCNA::TOMsimilarityFromExpr(
  not_grey_exprs,
  power = chosen_power)
colnames(dist_tom) <- not_grey_genes
rownames(dist_tom) <- colnames(dist_tom)

similarity_cluster <- flashClust::flashClust(as.dist(dist_tom), method = "average")

WGCNA::plotDendroAndColors(
  similarity_cluster,
  colors = initial_modules[["colors"]][not_grey_genes],
  "ME",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05,
  cex.colorLabels = 1.5,
  main = ("Cluster Dendrogram (WGCNA)"))

## In my initial pass, I did the clustering on the samples instead of genes and got
## two primary groups:
## 26,6,7,12,4,27,3,13,9,30,17,22,18,42,15,31,8,25,5,16,14,20,10,38
## 1,2,23,11,40,34,33,35,37,24,41,19,29,28,36,21,32,39
## I assume that these two groups will have some meaning vis a vis the monocyte samples?
first_group <- c(26,6,7,12,4,27,3,13,9,30,17,22,18,42,15,31,8,25,5,16,14,20,10,38)
second_group <- c(1,2,23,11,40,34,33,35,37,24,41,19,29,28,36,21,32,39)
unique(pData(l2input)[first_group, "tubelabelorigin"])
unique(pData(l2input)[second_group, "tubelabelorigin"])
## So, they are two distinct groups of donors...
table(pData(l2input)[first_group, "finaloutcome"])
table(pData(l2input)[second_group, "finaloutcome"])

table(pData(l2input)[first_group, "visitnumber"])
table(pData(l2input)[second_group, "visitnumber"])
```

```{r}
pander::pander(sessionInfo())
message("This is hpgltools commit: ", get_git_commit())
message("Saving to ", savefile)
# tmp <- sm(saveme(filename = savefile))
```
