---
title: "Setting up singularity images for TMRC analyses."
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
output:
  html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    keep_md: false
    mode: selfcontained
    number_sections: true
    self_contained: true
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  rmdformats::readthedown:
    code_download: true
    code_folding: show
    df_print: paged
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    width: 300
    keep_md: false
    mode: selfcontained
    toc_float: true
  BiocStyle::html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    keep_md: false
    mode: selfcontained
    toc_float: true
---

<style type="text/css">
body, td {
  font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
 font-size: 16px
}
body .main-container {
  max-width: 1600px;
}
</style>

```{r options, include=FALSE}
library(hpgltools)
library(reticulate)
tt <- try(devtools::load_all("~/hpgltools"))
knitr::opts_knit$set(
  progress = TRUE, verbose = TRUE, width = 90, echo = TRUE)
knitr::opts_chunk$set(
  error = TRUE, fig.width = 8, fig.height = 8, fig.retina = 2,
  fig.pos = "t", fig.align = "center", dpi = if (knitr::is_latex_output()) 72 else 300,
  out.width = "100%", dev = "png",
  dev.args = list(png = list(type = "cairo-png")))
old_options <- options(digits = 4,
                       stringsAsFactors = FALSE,
                       knitr.duplicate.label = "allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size = 12))
ver <- Sys.getenv("VERSION")
previous_file <- ""
ver <- format(Sys.Date(), "%Y%m%d")

##tmp <- sm(loadme(filename=paste0(gsub(pattern="\\.Rmd", replace="", x=previous_file), "-v", ver, ".rda.xz")))
rmd_file <- "tmrc3_singularity.Rmd"
```

# Introduction

It has been a while since I seriously considered making container
images for analyses. Let us fix that so that we can have a nice image
for the TMRC analyses.

# Singularity

Of the various options, I found singularity to be the most attractive.
I therefore wrote a default Makefile with a few targets to create and
manipulate images because I cannot be bothered to remember all of the
various commands.

# The images

I am hoping to create 2 image templates for the TMRC analyses:
preprocessing and analysis. I will therefore have a few locally
maintained shell scripts which contain the base image setup tasks, the
tasks required to setup the tools used, download the data, and perform
the work.

# Setting up the base image

I intend to use Debian stable.  I am copying the required setup files
into the current working directory and invoking make to create
the container.  I have a few targets which are intended to make this
easier to remember.

## Creating the base image

The following command runs singularity with options suitable to create
the image.  This to me is a little unnerving, because a bunch of stuff
gets run as root.

```{bash, eval=FALSE}
make tmrc3_analyses.simg
```

## Testing stuff out and making changes

The overlay target drops the user into the container with R/W
permissions.  It creates a directory 'tmrc3_analyses_overlay' which
may be modified at will.

```{bash, eval=FALSE}
make tmrc3_analyses.overlay
```

Once inside the container, one may use the 'setup_hpgltools.sh' script
to install a specific state of hpgltools and all of its various
prerequisites.  This takes ~ 45 minutes.

```{bash, eval=FALSE}
bin/setup_hpgltools.sh
```

```{r saveme}
pander::pander(sessionInfo())
message("This is hpgltools commit: ", get_git_commit())
this_save <- paste0(gsub(pattern = "\\.Rmd", replace = "", x = rmd_file), "-v", ver, ".rda.xz")
#message("Saving to ", this_save)
#tmp <- sm(saveme(filename = this_save))
```
